# Testing for time-varying extremal dependence

```{r changing-ext-dep-load-packages}
#| include: false
library(tidyverse)
library(tidyr)
library(magrittr)
library(SimCop)
library(pracma)
library(expm)
library(sn)
library(scales)
library(ggh4x)
library(CPAT)
library(maps)
library(patchwork)
library(ggpubr)
library(colorspace)
library(pbapply)
library(kableExtra)
library(reshape2)
library(e1071) # simulate Brownian bridge
library(tictoc)

options(dplyr.summarise.inform = FALSE)
options(knitr.kable.NA = "")
```

```{r changing-ext-dep-source-functions}
#| include: false
sapply(list.files(path = "R/changing-ext-dep", pattern = "*.R", full.names = TRUE, recursive = TRUE), source)
sapply(list.files(path = "R/general", pattern = "*.R", full.names = TRUE, recursive = TRUE), source)
```

Multivariate extreme value models typically assume that the observed data are independent observations of some fixed distribution. This requires that both the marginal distributions and the extremal dependence structure are constant throughout the observation period. As explained in Section XX with regards to univariate (marginal) modelling, the validity of this assumption may be called into question and non-stationary models are being developed to account for this. However, relatively little work has been done on non-stationarity in the extremal dependence structure, even though the same problems apply. Anthropogenic climate change is driving changes in the spatial structure of climate extremes [@zhouGlobalConcurrentClimate2023] and regulatory changes can cause structural changes in the joint tail behaviour of financial asset prices [@poonModellingExtremeValueDependence2003]. Thus, a crucial step in the modelling process is to determine whether it is reasonable to assume stationary dependence. In this chapter, we present a formal procedure for testing this assumption even in high dimensions.

At this point, we clarify an important distinction between *testing for* versus *modelling* non-stationary dependence. Both represent very challenging statistical problems: the underlying signal (e.g. climate change) may be very weak, perhaps only becoming apparent over very long observation periods; as usual inference is hampered by the inherent scarcity of relevant data. The latter task refers to the development of multivariate extreme value models that allow temporal non-stationarity in the dependence structure. For example, the regression model of @castro-camiloTimevaryingExtremeValue2018 and the spectral density ratio model of @decarvalhoSpectralDensityRatio2014 can incorporate covariate effects, including time. These models rely on parametric assumptions and are restricted to a small number of dimensions. To the best of our knowledge, the only existing work on *testing* for changing dependence is @dreesStatisticalInferenceChanging2023. Roughly speaking, their procedure involves partitioning the observation period into temporal blocks and testing for deviations in $\hat{H}$ between blocks. This is very computationally intensive and thus is restricted to $d\leq 5$ in practice. Our contribution is to devise a procedure that instead tests for deviations in $\hat{\Sigma}$, the empirical TPDM. Considering pairwise dependencies instead of the full angular measure eases the computational burden significantly and enables testing even in high dimensions. Our test achieves superior power in many realistic scenarios (Section XX). The trade-off is that neglecting higher-order dependencies necessarily incurs some information loss. As a result, our proposed method fails to detect certain types of dependence change (Section XX).  

## Framework and hypothesis

Suppose $\{\bm{X}(t) = (X_1(t),\ldots,X_d(t)):t\in[0,1]\}$ is an $\R_+^d$-valued, continuous time stochastic process with no serial dependence. Let $\|\cdot\|_2$ denote the Euclidean norm on $\R^d$. For $t\in[0,1]$, assume that the random vector $\bm{X}(t)$ is multivariate regularly varying (MRV) with index of regular variation $\alpha(t)=2$ and angular measure $H(\cdot\,; t)$ on $\mathbb{S}_+^{d-1}:=\{\bm{x}\in\R_+^d: \|\bm{x}\|_2=1\}$. Denoting by $R(t):=\|\bm{X}(t)\|_2$ and $\bm{\Theta}(t):=\bm{X}(t)/\|\bm{X}(t)\|_2$ the radial and angular components of $\bm{X}(t)$, respectively, the MRV property states that for all $z>0$ and Borel sets $\mathcal{B}\subset\mathbb{S}_+^{d-1}$,
\begin{equation}\label{eq-time-dependent-mrv}
    \lim_{u \to \infty} \frac{\mathbb{P}(R(t) > zu, \bm{\Theta}(t) \in \mathcal{B})}{\mathbb{P}(R(t) > u)} = z^{-\alpha(t)} H(\mathcal{B}; t).
\end{equation}
We assume $\bm{X}(t)$ is on Fréchet margins with shape parameter $\alpha(t)=2$, perhaps after a suitable marginal transformation. With this scaling, the angular measure of $\bm{X}(t)$ satisfies $H(\mathbb{S}_+^{d-1};t)=d$ for all $t\in[0,1]$. In the MRV paradigm, extremal dependence is fully characterised by the angular measure. Our working null and alternative hypotheses can be stated formally as
\begin{align}
\mathrm{H}_0 \, &: \,\, \forall t \in [0,1], \, H(\cdot \,; t) = H(\cdot \,; 1), \label{eq-changing-ext-dep-null-hyp} \\
\mathrm{H}_1 \, &: \,\, \exists t,\, H(\cdot \,; t) \neq H(\cdot \,; 1). \label{eq-changing-ext-dep-alt-hyp}
\end{align}
Our goal is to devise a statistical procedure for testing these hypotheses given a discretised sample path of $\{\bm{X}(t):t\in[0,1]\}$.

## Background and outlook

@dreesStatisticalInferenceChanging2023 tests \eqref{eq-changing-ext-dep-null-hyp} against \eqref{eq-changing-ext-dep-alt-hyp} via a large family $\mathcal{A}$ of subsets of $\mathbb{S}_+^{d-1}$ and suitably rescaled versions of stochastic processes
\begin{equation}\label{eq-drees-test-process}
\left\lbrace\int_0^t \hat{H}(A;s)\,\dee s - t \int_0^1 \hat{H}(A;s)\,\dee s : t\in[0,1]\right\rbrace, \qquad (A\in\mathcal{A}).
\end{equation}
Here $\hat{H}(A;s)$ denotes a non-parametric estimate of the angular measure $H(A;s)$ at time $s\in[0,1]$ -- see \eqref{eq-emp-angular-measure} for a formal definition. The null is rejected if any paths in \eqref{eq-drees-test-process} deviate from what would typically occur under the null. If $\mathcal{A}$ is sufficiently rich, then even very subtle dependence changes may be revealed, in principle. However, as the dimension $d$ increases the family of sets grows rapidly, typically $|\mathcal{A}|=\mathcal{O}(2^d)$. Consequently, the underlying computations become prohibitively intensive and the convergence $\hat{H(A;t)}\to H(A;t)$ of the non-parametric estimators is too slow. Thus, their method is primarily intended for the bivariate setting and is restricted to $d\leq 5$ in practice. Fundamentally, this limitation stems from the curse of dimensionality inherent to estimation of the angular measure. This impediment is exacerbated by the fact that inference must be performed *locally*, i.e. using only (extreme) observations lying within some small temporal neighbourhood.

Our approach mitigates this issue by concentrating on bivariate summaries of tail dependence instead of the full dependence structure. The $\mathcal{O}(d^2)$ coefficients of the TPDM encode second-order information about the local angular measure and can be more reliably estimated in high dimensions. The downside is that the TPDM contains incomplete information about the angular measure. This means our test is powerless in certain circumstances; a class of examples is provided in @sec-constant-tpdm.  

## The local (integrated) TPDM

Non-stationary dependence as in \eqref{eq-time-dependent-mrv} necessitates a time-dependent version of the TPDM. This is naturally defined via an integral with respect to the local angular measure. 

:::{#def-local-tpdm}
For $t\in[0,1]$, the local TPDM is the $d\times d$ matrix given by
\begin{equation}\label{eq-local-tpdm}
    \sigma_{ij}(t) = \int_{\mathbb{S}_+^{d-1}} \theta_i\theta_j \,\dee H(\bm{\theta};t), \qquad \Sigma(t)=(\sigma_{ij}(t))_{i,j=1,\ldots,d}.
\end{equation}
:::

The local TPDM summarises the tail dependence strength between pairs of components of $\bm{X}(t)$. Since $H(\cdot\,;t)$ is a valid angular measure, the local TPDM satisfies all the usual mathematical properties of a TPDM (@cooleyDecompositionsDependenceHighdimensional2019). 

While our principle objective is to detect changes in the local TPDM, it is common to devise statistical tests based on integrated versions of the quantity of interest. This strategy is employed by @dreesStatisticalInferenceChanging2023 -- consider \eqref{eq-drees-test-process}. This motivates the introduction of an integrated TDPM. 

:::{#def-integrated-tpdm}
For $t\in[0,1]$, the integrated TPDM is the $d\times d$ matrix given by
\begin{equation*}
    \psi_{ij}(t) = \int_0^t \sigma_{ij}(s) \,\dee s, \qquad \Psi(t)=(\psi_{ij}(t))_{i,j=1,\ldots,d}.
\end{equation*}
:::

The integrated TPDM is symmetric, positive semi-definite, and possesses the property that $\psi_{ij}(t)=0$ if and only if $X_i(s)$ and $X_j(s)$ are asymptotically independent for all $s\leq t$. Beyond this, it has no obvious interpretation. With standardised margins, we have that $\sigma_{ii}(t)=1$ and hence $\psi_{ii}(t)=t$ for all $i=1,\ldots,d$ and $t\in[0,1]$. The integrated TPDM can be equivalently defined via the so-called integrated angular measure of @dreesStatisticalInferenceChanging2023, since
\begin{equation*}
  \psi_{ij}(t) = \int_0^t  \int_{\mathbb{S}_+^{d-1}} \theta_i\theta_j \,\dee H(\bm{\theta};s) \,\dee s = \int_{\mathbb{S}_+^{d-1}} \theta_i\theta_j  \int_0^t \,\dee H(\bm{\theta};s) \,\dee s.
\end{equation*}
Due to this connection, many of the theoretical results contained in @dreesStatisticalInferenceChanging2023 transfer immediately to our methodology.

The matrices $\Sigma(t)$ and $\Psi(t)$ are symmetric with known diagonal entries, so nothing is lost by focussing exclusively on their strictly upper triangular elements. We introduce the following notation for referring to these entries. If $M=(m_{ij})$ denote an arbitrary $d\times d$ (random) matrix, the (random) vector obtained by row-wise vectorisation of its upper triangular elements shall be denoted by
\begin{equation*}
    \mathrm{vecu}(M) := (m_{12}, m_{13}, \ldots, m_{1d}, m_{23}, \ldots, m_{2d}, \ldots, m_{d-1,d}).
\end{equation*}
The components of $\bm{m}:=\mathrm{vecu}(M)$ are indexed according to the sub-indices of $M$, e.g. the first element of is $m_{12}$ rather than $m_1$. The upper-vectorised local TPDM and integrated TPDM are denoted by
\begin{align*}
  \bm{\sigma}(t) &:= \mathrm{vecu}(\Sigma(t)) = (\sigma_{12}(t), \sigma_{13}(t), \ldots, \sigma_{1d}(t), \sigma_{23}(t), \ldots, \sigma_{2d}(t), \ldots, \sigma_{d-1,d}(t)), \\
  \bm{\psi}(t) &:= \mathrm{vecu}(\Psi(t)) = (\psi_{12}(t), \psi_{13}(t), \ldots, \psi_{1d}(t), \psi_{23}(t), \ldots, \psi_{2d}(t), \ldots, \psi_{d-1,d}(t)).
\end{align*}
The dimension of these vectors is 
\begin{equation*}
\mathcal{D}:=|\{(i,j):1\leq i < j \leq d\}|=\binom{d}{2}=\frac{1}{2}d(d-1).
\end{equation*}
The following section concerns the estimation of these quantities.

## Inference

Suppose we observe a sample path of $\{\bm{X}(t):t\in[0,1]\}$ along $n$ discrete time-points according to an equidistant sampling scheme, yielding a collection of independent random vectors $\{\bm{X}(i/n):i=1,\ldots,n\}$. Our methodology could accommodate more general sampling schemes, but this one is the simplest and most commonly encountered. The general principle underlying the following is that extremal dependence at time $t\in[0,1]$ may be inferred from the $k$ most extreme observations lying within in a $h$-neighbourhood of $t$. The hyperparameters $h>0$ and $k\geq 1$ are called the *bandwidth* and *level*, respectively. Specifically, we define
\begin{equation*}
    \mathcal{I}(t) := \{i\in\{1,\ldots,n\} : i/n \in(t-h, t+h]\},
\end{equation*}
and among the observations $\{\bm{X}(i/n):i\in\mathcal{I}(t)\}$, only those whose norm exceeds a specified radial threshold will enter into our estimators. The threshold $\hat{u}(t)$ is set as the $k+1$ largest order statistic among $\{R(i/n):i \in \mathcal{I}(t)\}$; by construction, there will be exactly $k$ radial threshold exceedances. Selecting the level and bandwidth involves managing trade-offs between retaining an adequate number of samples (by increasing $h$ and $k$) while ensuring that estimation remains time-localised (reducing $h$) and free of bias due to observations from the distributional bulk (reducing $k$).

Our estimator for the local TPDM is founded on the empirical local angular measure defined in @dreesStatisticalInferenceChanging2023. For any $t\in[0,1]$, this random measure is given by
\begin{equation}\label{eq-emp-angular-measure}
\hat{H}(\cdot\,; t) := \frac{d}{k} \sum_{i\in\mathcal{I}(t)} \ind\{R(i/n)>\hat{u}(t),\bm{\Theta}(i/n)\in \cdot\}.
\end{equation}
Substituting \eqref{eq-emp-angular-measure} into \eqref{eq-local-tpdm} results in the following definition.

:::{#def-emp-local-tpdm}
For $t\in[0,1]$, the empirical local TPDM is the $d\times d$ matrix given by
\begin{equation}\label{eq-emp-local-tpdm}
\hat{\sigma}_{ij}(t) := \int_{\mathbb{S}_+^{d-1}} \theta_i\theta_j \,\dee \hat{H}(\bm{\theta}; t) = \frac{d}{k} \sum_{l\in\mathcal{I}(t)} \Theta_i(l/n)\Theta_j(l/n) \ind\{R(l/n)>\hat{u}(t)\}, \qquad \hat{\Sigma}(t)=(\hat{\sigma}_{ij}(t)).
\end{equation}
:::

We recognise \eqref{eq-emp-local-tpdm} as simply a time-localised version of the familiar empirical TPDM (Equation 5 in @cooleyDecompositionsDependenceHighdimensional2019). Thus it retains all the usual properties of an empirical TPDM.

Estimating the integrated TPDM is slightly more complicated, because $\Psi(t)$ depends on the full history of $\Sigma(s)$ over the continuous interval $s\in[0,t]$. This is achieved by the same block-based construction used by @dreesStatisticalInferenceChanging2023. First, we partition the full observation period $[0,1]$ into blocks of width $2h$; each block contains $b:=2nh$ observations. (Henceforth, assume that the number of blocks $t/(2h)=n/b$ is an integer.) Next, we estimate the local TPDM at at the block centres $t\in\{h, 3h, \ldots,(2n/b -1)h\}$ using \eqref{eq-emp-local-tpdm} with the bandwidth in $\mathcal{I}(t)$ set equal to half the block width (that is, $h$). Then, we interpolate the local TPDM estimates according to an assumption of constant dependence within each block, so that for any index pair $1\leq i < j \leq d$, $\hat{\sigma}_{ij}(s)$ constitutes a piecewise constant function of $s$ on $[0,1]$. The entries of the empirical integrated TPDM are given by the corresponding time-integrals of these functions.

:::{#def-emp-integrated-tpdm}
For $t\in[0,1]$, the empirical integrated TPDM is the $d\times d$ matrix given by
\begin{equation}\label{eq-emp-integrated-tpdm-1}
\hat{\psi}_{ij}(t) := \int_0^t \hat{\sigma}_{ij}((2 \lceil s/(2h) \rceil - 1)h) \,\dee s, \qquad \hat{\Psi}(t)=(\hat{\psi}_{ij}(t)).
\end{equation}
:::

This can be equivalently and more conveniently expressed as
\begin{equation}\label{eq-emp-integrated-tpdm-2}
\hat{\Psi}(t) := 2h \sum_{l=1}^{L(t)} \hat{\Sigma}(s_l) + (t - 2h L(t)) \hat{\Sigma}(s_{L(t) +1}), \qquad L(t):=\lfloor t/(2h)\rfloor,\qquad s_l:= (2l-1)h.
\end{equation}
The first term in \eqref{eq-emp-integrated-tpdm-2} corresponds to the $L(t)$ whole blocks in $[0,t]$, each of which receive a full weighting of $2h$. The second term receives a reduces weight since it relates to the partial block containing $t$.  

While the formulation \eqref{eq-emp-integrated-tpdm-1} looks less cumbersome, \eqref{eq-emp-integrated-tpdm-2} will prove more convenient, both computationally and mathematically. In particular, it reveals that the $\hat{\psi}_{ij}(t)$ is a weighted sum of the independent random variables $\sigma_{ij}(s_1),\ldots,\sigma_{ij}(s_{L(t)+1}$. Independence is due to the blocks being non-overlapping and is crucial in the elicitation of the asymptotic results to follow. 

### Asymptotic theory

We now formulate the asymptotic properties of the estimators $\hat{\bm{\sigma}}(t):=\mathrm{vecu}(\hat{\Sigma}(t))$ and $\hat{\bm{\psi}}(t):=\mathrm{vecu}(\hat{\Psi}(t))$. Henceforth, the bandwidth and level are sequences satisfying $h\to 0$, $nh\to\infty$, $k\to\infty$, and $k/(nh)\to 0$ as $n\to\infty$.

The first result concerns asymptotic normality of the empirical local TPDM. As remarked earlier, $\hat{\Sigma}(t)$ is simply an empirical TPDM based on the data subset $\{\bm{X}(i/n):i/n\in(t-h,t+h]\}$, which comprises $2nh$ observations. Asymptotically, by assumption, the size of this restricted sample $2nh\to\infty$, the number of observations entering the estimator $k\to\infty$, and the proportion of observations entering the estimator $k/(2nh)\to 0$. Thus all the conditions required for asymptotic normality of the empirical TPDM hold (see @larssonExtremalDependenceMeasure2012 and Section 6.1 in @leePartialTailCorrelation2023).

:::{#prp-asymptotic-emp-local-tpdm}
For any $t\in[0,1]$,
\begin{equation}\label{eq-emp-local-tpdm-asymptotic}
    k^{1/2}(\hat{\bm{\sigma}}(t) - \bm{\sigma}(t)) \to N(\bm{0}, V(t))
\end{equation}
as $n\to\infty$. The $\mathcal{D}\times\mathcal{D}$ asymptotic covariance matrix is given by
\begin{equation}
  V(t):=\mathrm{Cov}(\mathrm{vecu}(d\tilde{\bm{\Theta}}(t)\tilde{\bm{\Theta}}(t)^T)), \qquad \tilde{\bm{\Theta}}(t)\sim d^{-1}H(\cdot \,; t).
\end{equation}
:::

The diagonal entries $V_{ij,ij}(t)$ of $V(t)$ relate to the asymptotic variance of the estimators $\hat{\sigma}_{ij}(t)$. The off-diagonal entries $V_{ij,lm}(t)$ relate to the asymptotic covariance between $\hat{\sigma}_{ij}(t)$ and $\hat{\sigma}_{lm}(t)$. Ordinarily $V(t)$ is unknown but will be present as a nuisance parameter in our test statistics. For now we assume that $V(t)$ is known; later it will be replaced by a plug-in estimator. 

Considering \eqref{eq-emp-integrated-tpdm-2} and @prp-asymptotic-emp-local-tpdm, the components of $\bm{\psi}(t)$ are weighted sums of independent, asymptotically normal random variables. By a functional central limit theorem type argument it follows that, with suitable rescaling, the stochastic process $\{\hat{\psi}_{ij}(t):t\in[0,1]\}$ converges in distribution to a Gaussian processes. 

:::{#prp-asymptotic-emp-local-integrated-tpdm}
The $\mathcal{D}$-dimensional, continuous-time stochastic process
  \begin{equation}\label{eq-prelim-test-process}
    \left\lbrace \left(\frac{k}{h}\right)^{1/2} \left( \hat{\bm{\psi}}(t) - \bm{\psi}(t) \right) : t\in[0,1] \right\rbrace,
  \end{equation}
converges to the $\mathcal{D}$-dimensional centred Gaussian process $\{\bm{Y}(t):t\in[0,1]\}$ with covariance function
\begin{equation}\label{eq-prelim-test-process-cov-function}
    \mathrm{Cov}(Y_{ij}(s),Y_{lm}(t)) = 2 \int_0^{\min(s,t)} V_{ij,lm}(\tau)\,\dee \tau.
\end{equation}
:::

::: {.proof}
Write proof here.
:::

The drift and diffusion coefficients associated with each univariate process $\{Y_{ij}(t):t\in[0,1]\}$ are controlled by underlying integrated TPDM $\{\psi_{ij}(t):t\in[0,1]\}$ and asymptotic variance process $\{V_{ij,ij}(t):t\in[0,1]\}$, respectively. Meanwhile, the cross-correlation between $\{Y_{ij}(t):t\in[0,1]\}$ and $\{Y_{lm}(t):t\in[0,1]\}$ is determined by the asymptotic covariance $\{V_{ij,lm}(t):t\in[0,1]\}$.

Under the null hypothesis \eqref{eq-changing-ext-dep-null-hyp}, the asymptotic variance-covariance matrix $V=V(t)$ is independent of time and the covariance function \eqref{eq-prelim-test-process-cov-function} simplifies to $\mathrm{Cov}(\bm{Y}(s),\bm{Y}(t)) = 2V\min(s,t)$. Thus, upon pre-multiplying \eqref{eq-prelim-test-process} by $(2V)^{-1/2}$, the distribution of the limiting process equals that of a standard $\mathcal{D}$-dimensional standard Brownian motion.

### Hypothesis testing

In view of @prp-asymptotic-emp-local-integrated-tpdm and the ensuant discussion, we define a $\mathcal{D}$-dimensional test process $\{\hat{\bm{Z}}(t):t\in[0,1]\}$ as
\begin{equation}\label{eq-test-process}
  \hat{\bm{Z}}(t) := \left(\frac{k}{2 h}\right)^{1/2} V(t)^{-1/2}(\hat{\bm{\psi}}(t) - t\hat{\bm{\psi}}(1)).
\end{equation}
The nuisance parameter $V(t)$ standardises the processes $\hat{Z}_{ij}(t)$ and removes cross-correlation between them. Its inclusion is vital for ensuring a convenient asymptotic null distribution for our test statistics and thus allowing critical values to be readily available without recourse to simulation. Generally, $V(t)$ may be assumed to be invertible, since the off-diagonal TPDM entries are not constrained to equal any particular value. This would not be the case had the vectorised quantities $\bm{\sigma}(t),\bm{\psi}(t)$ included components pairs $i=j$: the diagonal entries of the TPDM satisfy $\mathrm{trace}(\Sigma(t))=\sigma_{11}(t)+\ldots+\sigma_{dd}(t)=d$ for all $t\in[0,1]$, forming a linear combination of components with zero variance. 

From the test process we define Kolmogorov-Smirnov (KS) and Cramér-von-Mises (CM) type test statistics by
\begin{align}
    T^{(KS)} &:= \sup_{t\in[0,1]}\left\lVert \hat{\bm{Z}}(t) \right\rVert_\infty, \label{eq-test-stat-ks} \\
    T^{(CM)} &:= \sup_{1\leq i<j\leq d} \left\lVert \hat{Z}_{ij}(t) \right\rVert_{L^2[0,1]}^2, \label{eq-test-stat-cm}
\end{align}
where $\|\bm{x}\|_\infty := \max\{ |x_i|:i=1,\ldots,\mathcal{D}\}$ denotes the sup-norm in $\R^\mathcal{D}$ and $\|Y(t)\|_{L^2[0,1]}^2:=\int_0^1 |Y(t)|^2\,\dee t$ denotes the $L^2$-norm of a stochastic process on $[0,1]$. Their asymptotic null distributions are given below.

:::{#prp-asymptotic-test-stat}
Under the null hypothesis \eqref{eq-changing-ext-dep-null-hyp},
\begin{equation}\label{eq-asymptotic-test-stat}
    T^{(KS)} \to \sup_{t\in[0,1]} \|\bm{B}(t)\|_\infty \overset{d}{=} \sup_{1\leq i<j\leq d} K_{ij}, \qquad 
    T^{(CM)} \to \sup_{1\leq i<j\leq d} \|B_{ij}(t)\|_{L^2[0,1]}^2,
\end{equation}
where $\bm{B}(t)=(B_{ij}(t) : 1\leq i < j \leq d)$ denotes a standard $\mathcal{D}$-dimensional Brownian bridge and $\{K_{ij}:1\leq i<j\leq d\}$ is a collection of $\mathcal{D}$ independent Kolmogorov random variables. 
:::

::: {.proof}
Under the null hypothesis, $\bm{\psi}(t)=t\cdot \bm{\psi}(1)$ and therefore
\begin{align*}
    \hat{\bm{Z}}(t)
    &=  (2V)^{-1/2} \left(\frac{k}{h}\right)^{1/2} \left(\hat{\bm{\psi}}(t) - \bm{\psi}(t) - t(\hat{\bm{\psi}}(1) - \bm{\psi}(1))\right) \\
    &\to (2V)^{-1/2} (\bm{Y}(t) - t \bm{Y}(1)) \\
    &\overset{d}{=} \bm{W}(t) - t\bm{W}(1) \\
    &\overset{d}{=} \bm{B}(t),
\end{align*}
where $\bm{W}(t)=(W_{ij}(t) : 1\leq i < j \leq d)$ denotes a standard $\mathcal{D}$-dimensional Brownian motion. The independent random variables $K_{ij}:=\sup_{t\in[0,1]} |B_{ij}(t)|$, for $1\leq i<j\leq d$, are Kolmogorov distributed by definition. 
:::

Denoting the Kolmogorov distribution function by $F_K$,
\begin{equation}\label{eq-ks-test}
\ind\{T^{(KS)}>c_\alpha\}, \qquad c_{\alpha} = F_K^{-1}((1-\alpha)^{1/\mathcal{D}})
\end{equation}
constitutes an asymptotic level $\alpha$ test. The critical value $c_\alpha$ represents the value for which the probability that a set of $\mathcal{D}$ independent one-dimensional Brownian bridges all remain in the region $(-c_{\alpha},c_{\alpha})$ equals $1-\alpha$. A CM-type test is constructed analogously, except the distribution of the $L^2$-norm of a Brownian bridge is unknown, so the critical values must be obtained via simulation. To this end, we generate 50,000 Brownian bridge sample paths on a fine mesh, compute the appropriate $L^2$ norms via numerical integration, and obtain critical values by estimating the various quantiles of interest. Critical values for selected dimensions and significance levels are listed in @tbl-critical-values.

```{r make-tbl-critical-values}
#| label: tbl-critical-values
#| tbl-cap: "Asymptotic critical values for selected dimensions and significance levels."
#| tbl-subcap: "Critical values for selected dimensions and significance levels."

bb_L2 <- readRDS("scripts/changing-ext-dep/results/bb_L2.RDS")

expand_grid(d = c(2, 3, 4, 5, 10, 15, 20, 25),
            alpha = c(0.01, 0.05, 0.1),
            type = c("CM", "KS")) %>%
  mutate(D = d * (d - 1) / 2) %>%
  relocate(D, .after = d) %>%
  mutate(crit_val = case_when(
    type == "CM" ~ quantile(bb_L2, probs = (1 - alpha)^(1 / D)),
    type == "KS" ~ CPAT:::qkolmogorov((1 - alpha)^(1 / D)))) %>%
  arrange(d, alpha) %>%
  pivot_wider(names_from = c(alpha, type), values_from = crit_val) %>%
  kbl(col.names = c("$d$", "$\\mathcal{D}$", rep(c("CM", "KS"), 3)),
      booktabs = TRUE, linesep = "", digits = 3, escape = FALSE) %>%
  add_header_above(c(' ' = 2, '$\\\\alpha=0.01$' = 2, '$\\\\alpha=0.05$' = 2, '$\\\\alpha=0.10$' = 2), line_sep = 1, escape = FALSE)
```

It remains to explain how we deal with the nuisance parameter(s) $\{V(t):t\in[0,1]\}$. Our approach is simply to estimate it from the data. There are various ways this could be done, but we find the following works well in practice. Under the null, the (single) nuisance parameter $V=V(t)$ represents the covariance matrix of $\mathrm{vecu}(d\bm{\Theta}\bm{\Theta}^T)$, where the redundant time-dependence in $\bm{\Theta}=\bm{\Theta}(t)$ is suppressed. Our estimator for $V$ will be the associated sample covariance matrix estimated from the entire set of $k_{\mathrm{total}}:=kn/b$ radial threshold exceedances taken from all blocks. That is
\begin{align*}
    \hat{V} &:= \frac{1}{k_{\mathrm{total}}} \sum_{l=1}^n W_l W_l^T \ind\{R(l/n)>\hat{u}((2\lceil l/b \rceil - 1)h)\} \\
    W_l &:= \mathrm{vecu}(d\bm{\Theta}(l/n)\bm{\Theta}(l/n)^T) - \hat{\bm{\sigma}}((2\lceil l/b \rceil - 1)h).
\end{align*}
Provided the *rank condition* $k_{\mathrm{total}}>\mathcal{D}$ is satisfied, the estimator $\hat{V}$ is full-rank and therefore invertible. For a fixed sample size and set of tuning parameters, the rank condition imposes an upper limit on the dimension, roughly $d<\sqrt{2k_{\mathrm{total}}}$. It seems natural that such a restriction should exist: reliable inference in high-dimensional settings requires commensurate data. For fixed $n$, we may reduce $b$ and/or increase $k$ in order to enlarge the effective sample size, but these parameters are subject to their own particular trade-offs that will influence the performance of the test. Alternatively, one could substitute $V^{-1}$ with the pseudoinverse to circumvent invertibility concerns. This avenue is not explored and in any case it doesn't seem sensible to proceed with the test in circumstances where violation of the rank condition indicates there is insufficient data for the task at hand. 

## Simulation experiments

In this section, we present a series of numerical experiments demonstrating our method's performance and, where applicable, draw conclusions regarding its relative merits compared to @dreesStatisticalInferenceChanging2023.

### Data generating processes

Suppose $\bm{X}(t)$ has dimension $d$ and its extremal dependence structure is parametrised by $\vartheta(t)\in\Omega$, where $\Omega$ is a convex parameter space. Let $\vartheta_0,\vartheta_1\in\Omega$ denote arbitrary parameters. We consider three scenarios for how the dependence of $\bm{X}(t)$ varies over time:

1. **Constant:** the parameter is fixed, i.e. $\vartheta(t)=\vartheta_0$.
2. **Jump:** the parameter changes (instantaneously) from $\vartheta_0$ to $\vartheta_1$ at a change point $\tau\in(0,1)$, i.e. $\vartheta(t)=\vartheta_0 \ind\{t <\tau\} + \vartheta_1 \ind\{t \geq \tau\}$. In all experiments we set $\tau=0.5$.
3. **Linear:** the parameter evolves linearly from $\vartheta_0$ to $\vartheta_1$, i.e. $\vartheta(t)=\vartheta_0 + t(\vartheta_1 - \vartheta_0)$. Convexity of $\Omega$ guarantees that $\vartheta(t)\in\Omega$ for all $t\in[0,1]$.

The parametric models we consider are as follows:

1. **Symmetric logistic (SL):** the dependence structure is characterised via the extreme value copula given by
    \begin{equation*}
    C(u_1,\ldots,u_d) = \exp\left( - \left[\sum_{j=1}^d (-\log u_j)^{\vartheta(t)} \right]^{1/\vartheta(t)} \right).
    \end{equation*}
    The parameter space is $\Omega=[1,\infty)$, with asymptotic independence when $\vartheta(t)=1$ and complete asymptotic dependence as $\vartheta(t)\to \infty$.
2. **Hüsler-Reiss (HR):** the dependence structure is characterised by the variogram $\Gamma(t)=\vartheta(t)\Gamma_0$, where $\Gamma_0\in\R^{d\times d}$ is a conditionally negative definite matrix and $\vartheta(t)\in\Omega=(0,\infty)$. Under this model, the extremal dependence coefficient between $X_i$ and $X_j$ at time $t\in [0,1]$ is $\chi_{ij}(t)=2\bar{\Phi}(\sqrt{\Gamma_{ij}(t)}/2)$, where $\bar{\Phi}$ is the survival function of the standard normal distribution. Asymptotic independence between $X_i$ and $X_j$ occurs as $\Gamma_{ij}(t)\to\infty$ and complete asymptotic dependence occurs if $\Gamma_{ij}(t)=0$. The multiplicative scalar $\vartheta(t)$ has the effect of increasing ($0<\vartheta(t)<1$) or decreasing ($\vartheta(t)>1$) the strength of all pairwise dependencies (relative to $\Gamma_0$). While not strictly necessary, we take $\vartheta_0=1$ so that $\Gamma_0$ parametrises the dependence at time $t=0$. For fixed $d$, the elements of the initial variogram $\Gamma_0$ are generated randomly using (elements of) the procedure outlined in Appendix B1 in @fomichovSphericalClusteringDetection2023. Specifically, we set $\Gamma_{0,ij}=\frac{3}{d}\|\bm{h}_i-\bm{h}_j\|_2^2$, where $\bm{h}_1,\ldots,\bm{h}_d$ are independent $d$-dimensional random vectors whose components are independent, identically distributed Pareto random variables with shape parameter equal to 2.5. The scaling factor $3/d$ ensures a suitable distribution for the extremal dependence coefficients.

Data are generated via the \texttt{rmev} function in the \texttt{mev} package. Our nomenclature for  referring to the six qualitatively different models is as follows: HR-jump refers to the Hüsler-Reiss model with a jump change in dependence, SL-linear refers to the symmetric logistic model with linearly evolving dependence and so on. 

For bivariate experiments @dreesStatisticalInferenceChanging2023 is included as a comparator. Results pertaining to their test are based on $\mathcal{A}=\{A_y:y=0.01,0.02,\ldots,0.99\}$, where $A_y :=\{\bm{\theta}\in\mathbb{S}_+^1: \theta_1 \leq y\}\subset \mathbb{S}_+^1$. 

### Results: asymptotic (large sample) performance

In an idealised setting with infinite data, the asymptotic theory in the previous section holds exactly. In practice we are naturally limited to finite samples, but we can validate our theoretical results empirically by taking $n$ sufficiently large and choosing $k$ and $b$ are appropriately. Specifically, the test's p-values under \eqref{eq-asymptotic-test-stat} should be uniformly distributed under the null and the test's empirical power under fixed alternatives should converge to 100\%.

First, we examine the asymptotic empirical distribution of the test statistics under the null. We generate 350 samples of size $n=10^6$ from the SL-constant ($\vartheta_0=2$) and HR-constant ($\vartheta_0=1$) models in dimensions $d\in\{2,5\}$. The bandwidth is $h=10^{-3}$ and the level is $k=50$. This yields 500 blocks of size $b=2,000$, a sampling fraction $k/b=2.5\%$, and an overall effective sample size of $k_{\mathrm{total}}=25,000$. @fig-qqplot-null depicts the empirical quantile functions of the p-values (upper plots) and test statistics (lower plots) against their theoretical counterparts. For the KS-type test, the theoretical quantiles in the QQ plots are computed using the exact Kolmogorov quantile function; for the CM-type test they are estimated from the set of simulated Brownian bridges discussed earlier. For both dimensions and models, the empirical p-values are approximately uniformly distributed. This indicates that for all nominal sizes the corresponding tests will approximately maintain the desired level. Analogous plots for @dreesStatisticalInferenceChanging2023 method can be found in Figure 7 within their Supplementary Material. 

```{r load-sim-results}
sim_results <- list(file.path("scripts", "changing-ext-dep", "results", "sim-test-results-seed1.RDS"),
                    file.path("scripts", "changing-ext-dep", "results", "sim-test-results-seed2.RDS"),
                    file.path("scripts", "changing-ext-dep", "results", "sim-test-results-seed3.RDS"),
                    file.path("scripts", "changing-ext-dep", "results", "sim-test-results-seed4.RDS"),
                    file.path("scripts", "changing-ext-dep", "results", "sim-test-results-seed5.RDS"),
                    file.path("scripts", "changing-ext-dep", "results", "sim-test-results-seed6.RDS"),
                    file.path("scripts", "changing-ext-dep", "results", "sim-test-results-seed7.RDS"),
                    file.path("scripts", "changing-ext-dep", "results", "sim-test-results-seed8.RDS")) %>%
  lapply(readRDS) %>%
  bind_rows()
```

```{r make-fig-qqplot-null}
#| label: fig-qqplot-null
#| fig-cap: "Large sample QQ plots for the p-values (top) and test statistics (bottom) associated with the KS test (left) and CM test (right). Based on 350 simulations from the SL- and HR-constant models with $n=10^6$, $b=2000$ and $k=50$."
#| fig-scap: "Large sample QQ plots for the KS/CM p-values and test statistics."
#| fig-height: 4

q_seq <- seq(from = 0, to = 1, by = 0.001)

p1 <- sim_results %>%
  filter(change_type == "none", n == 10^6, test_method == "pawley", test_type == "ks") %>%
  ggplot(aes(sample = 1 - CPAT:::pkolmogorov(test_stat)^(d*(d-1)/2), colour = model, linetype = as.factor(d))) +
  stat_qq(distribution = qunif, geom = "path") +
  geom_abline(intercept = 0, slope = 1, linetype = 1, colour = "darkgrey") +
  xlab("Uniform quantiles") +
  ylab("Sample quantiles") +
  ggtitle("KS p-values") +
  labs(colour = "Model", linetype = "Dimension, d") +
  scale_x_continuous(limits = c(0, 1), expand = c(0, 0), breaks = breaks_pretty(n = 3)) +
  scale_y_continuous(limits = c(0, 1), expand = c(0, 0), breaks = breaks_pretty(n = 3)) +
  scale_color_manual(labels = c("HR", "SL"), values = c("red", "blue")) +
  scale_linetype_manual(labels = c("2", "5"), values = c(1, 2)) +
  theme_light() +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        plot.title = element_text(hjust = 0.5, face = "bold", size = 10))

p2 <- sim_results %>%
  filter(change_type == "none", n == 10^6, test_method == "pawley", test_type == "cm") %>%
  rowwise() %>%
  mutate(p_value = 1 - mean(bb_L2 < test_stat)^(d*(d-1)/2)) %>%
  ungroup() %>%
  group_by(model, d) %>%
  summarise(q_p_value = list(quantile(p_value, q_seq))) %>%
  ungroup() %>%
  unnest_longer(q_p_value, values_to = "quantile_p_value", indices_to = "prob") %>%
  mutate(prob = as.numeric(gsub("%", "", prob)) / 100) %>%
  ggplot(aes(x = prob, y = quantile_p_value, colour = model, linetype = as.factor(d))) +
  geom_path() +
  geom_abline(intercept = 0, slope = 1, linetype = 1, colour = "darkgrey") +
  xlab("Uniform quantiles") +
  ylab("Sample quantiles") +
  ggtitle("CM p-values") +
  labs(colour = "Model", linetype = "Dimension, d") +
  scale_x_continuous(limits = c(0, 1), expand = c(0, 0), breaks = breaks_pretty(n = 3)) +
  scale_y_continuous(limits = c(0, 1), expand = c(0, 0), breaks = breaks_pretty(n = 3)) +
  scale_color_manual(labels = c("HR", "SL"), values = c("red", "blue")) +
  scale_linetype_manual(labels = c("2", "5"), values = c(1, 2)) +
  theme_light() +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        plot.title = element_text(hjust = 0.5, face = "bold", size = 10))

p3 <- sim_results %>%
  filter(change_type == "none", n == 10^6, d == 2, test_method == "pawley", test_type == "ks") %>%
  ggplot(aes(sample = test_stat, colour = model, linetype = as.factor(d))) +
  stat_qq(distribution = CPAT:::qkolmogorov, geom = "path") +
  geom_abline(intercept = 0, slope = 1, linetype = 1, colour = "darkgrey") +
  xlab("Kolmogorov quantiles") +
  ylab("Sample quantiles") +
  ggtitle("KS test statistics") +
  labs(colour = "Model", linetype = "Dimension, d") +
  scale_x_continuous(breaks = breaks_pretty(n = 3)) +
  scale_y_continuous(breaks = breaks_pretty(n = 3)) +
  scale_color_manual(labels = c("HR", "SL"), values = c("red", "blue")) +
  scale_linetype_manual(labels = c("2", "5"), values = c(1, 2)) +
  theme_light() +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        plot.title = element_text(hjust = 0.5, face = "bold", size = 10))

p4 <- sim_results %>%
  filter(change_type == "none", n == 10^6, d == 2, test_method == "pawley", test_type == "cm") %>%
  group_by(model, d) %>%
  summarise(q_test_stat = list(quantile(test_stat, q_seq))) %>%
  ungroup() %>%
  unnest_longer(q_test_stat, values_to = "quantile_test_stat", indices_to = "prob") %>%
  mutate(prob = as.numeric(gsub("%", "", prob)) / 100) %>%
  mutate(quantile_bb_L2 = quantile(bb_L2, probs = prob)) %>%
  ggplot(aes(x = quantile_bb_L2, y = quantile_test_stat, colour = model, linetype = as.factor(d))) +
  geom_path() +
  geom_abline(intercept = 0, slope = 1, linetype = 1, colour = "darkgrey") +
  xlab("Theoretical quantiles (simulated)") +
  ylab("Sample quantiles") +
  ggtitle("CM test statistics") +
  labs(colour = "Model", linetype = "Dimension, d") +
  scale_x_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.03)), breaks = breaks_pretty(n = 3)) +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.03)), breaks = breaks_pretty(n = 3)) +
  scale_color_manual(labels = c("HR", "SL"), values = c("red", "blue")) +
  scale_linetype_manual(labels = c("2", "5"), values = c(1, 2)) +
  theme_light() +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        plot.title = element_text(hjust = 0.5, face = "bold", size = 10))

ggarrange(p1, p2, p3, p4, ncol = 2, nrow = 2, common.legend = TRUE, legend = "right")
```

Next, we check that our procedure can leverage abundant information to detect dependence changes with high probability (i.e. is consistent under certain fixed alternatives). The experimental procedure is unchanged, except that data are now generated from SL-jump ($\vartheta_0=2$, $\vartheta_1=2.5$) and HR-jump ($\vartheta_0=1$, $\vartheta_1=1.5$) models. These values are chosen to bring about relatively subtle shifts in the dependence structure. Nevertheless our method consistently and overwhelmingly identifies non-stationary dependence. For the SL-jump data all p-values equal zero, up to numerical precision. For HR-jump data, the 90\% empirical quantile of the p-values is $2\times 10^{-6}$.

```{r}
#| eval: false
# this code gives the figures quoted above
sim_results %>%
  filter(change_type == "jump", n == 10^6, test_method == "pawley") %>% 
  rowwise() %>%
  mutate(p_value = case_when(
    test_type == "ks" ~ 1 - CPAT:::pkolmogorov(test_stat)^(d*(d-1)/2),
    .default = 1 - mean(bb_L2 < test_stat)^(d*(d-1)/2))) %>% 
  ungroup() %>%
  group_by(model, as.factor(d), test_type) %>%
  summarise(max_p_values = max(p_value))
```

### Results: finite sample performance

In finite sample settings, empirical size of an asymptotic test will generally differ from the nominal size; we only guarantee that the correct level is attained asymptotically. The hope is that convergence occurs with sufficient rapidity that this difference is acceptably small. We conduct repeated simulations from the SL-constant ($\vartheta_0=2$) and HR-constant ($\vartheta_0=1$) models in dimensions $d\in\{2, 5, 10, 25\}$ with sample sizes $n\in\{2500,5000,10000\}$. For each data set, we apply hypothesis tests with nominal level 5\%, based on various combinations of hyperparameters $b$ and $k$. Specifically, the number of blocks is $n/b\in\{25,50\}$ and the proportion of extreme observations within each block is $k/b\in\{0.05, 0.10, 0.15\}$. @tbl-type1 reports the empirical Type I error rates of these tests. Blank cells indicate that the corresponding tuning parameters violate the rank condition or the condition $k\leq d$. Large sample results are included in the tables for completeness -- recall that these are only available in dimensions $d\in\{2,5\}$. 

The size of our test exceeds the nominal level by at most 1.4\% and 3.3\% for the SL and HR models, respectively. Moreover, and arguably more pertinently, under any scenario (i.e. model, sample size and dimension) there exists hyperparameters for which this difference is at most 0.6\%. This suggests that where there is a large discrepancy between the nominal and empirical error rates, suboptimal hyperparameter selection may be the proximate cause. Having said that, the general stability in the empirical error rates demonstrates a certain degree of robustness to hyperparameter choices. The KS-based test is universally more conservative than the CM-based test, particularly for larger block lengths. The same pattern is observed in @dreesStatisticalInferenceChanging2023; it stems from the fact that a coarsely discretised path may only attain its supremum at a small number of points, whereas the corresponding critical values arise from suprema of continuous processes. 

```{r make-tbl-type1}
#| label: tbl-type1
#| tbl-cap: "Empirical Type I error rates (%) across repeated simulations. The number of simulations is $N=1000$ if $n\\leq 10^4$ and $d\\leq 5$, or $N=300$ otherwise. All tests have nominal size 5%."
#| tbl-subcap:
#|   - "SL-constant."
#|   - "HR-constant."
#| tbl-scap: "Empirical Type I error rates."
#| layout: "[[1], [-1], [1]]"

sim_results %>%
  filter(change_type == "none", model == "log") %>%
  group_by(n, d, n_blocks, k_frac, test_method, test_type) %>%
  summarise(empirical_type1 = 100 * mean(reject_H0)) %>%
  arrange(n, d, n_blocks, k_frac) %>%
  pivot_wider(names_from = c(d, test_method, test_type), values_from = empirical_type1) %>%
  kbl(col.names = c("$n$", "$n/b$", "$k/b$", rep(c("CM", "KS"), 5)),
      booktabs = TRUE, digits = c(0, 0, 3, rep(1, 10)), format.args = list(big.mark = ",",
  scientific = FALSE), escape = FALSE) %>%
  add_header_above(c(' ' = 3, 'Drees' = 2, 'Pawley' = 2, 'Pawley' = 2, 'Pawley' = 2, 'Pawley' = 2), line_sep = 1) %>%
  add_header_above(c(' ' = 3, "$d=2$" = 4, "$d=5$" = 2, "$d=10$" = 2, "$d=25$" = 2), line_sep = 1, bold = TRUE, escape = FALSE) %>%
  collapse_rows(columns = 1:2, latex_hline = "major", row_group_label_position = "first") %>%
  kable_styling(latex_options = "striped", font_size = 9)
# 
sim_results %>%
  filter(change_type == "none", model == "hr") %>%
  group_by(n, d, n_blocks, k_frac, test_method, test_type) %>%
  summarise(empirical_type1 = 100 * mean(reject_H0)) %>%
  arrange(n, d, n_blocks, k_frac) %>%
  pivot_wider(names_from = c(d, test_method, test_type), values_from = empirical_type1) %>%
  kbl(col.names = c("$n$", "$n/b$", "$k/b$", rep(c("CM", "KS"), 5)),
      booktabs = TRUE, digits = c(0, 0, 3, rep(1, 10)), format.args = list(big.mark = ",",
  scientific = FALSE), escape = FALSE) %>%
  add_header_above(c(' ' = 3, 'Drees' = 2, 'Pawley' = 2, 'Pawley' = 2, 'Pawley' = 2, 'Pawley' = 2), line_sep = 1) %>%
  add_header_above(c(' ' = 3, "$d=2$" = 4, "$d=5$" = 2, "$d=10$" = 2, "$d=25$" = 2), line_sep = 1, bold = TRUE, escape = FALSE) %>%
  collapse_rows(columns = 1:2, latex_hline = "major", row_group_label_position = "first") %>%
  kable_styling(latex_options = "striped", font_size = 9)
```

Next we examine the empirical power under alternatives. @fig-bivariate-power shows the power across a range of scenarios where data generating process undergoes a jump/linear dependence change of varying magnitude. All values are based on 1000 bivariate datasets of size $n=2500$; the six panels within each sub-plot illustrate the power for various hyperparameter choices. The nominal size of the tests (5\%) is indicated by the grey dashed line. Of course, when the null hypothesis is true ($\vartheta_1=2$ for SL, $\vartheta_1=1$ for HR), the power reverts to approximately this level. 

The power doesn't appear to be overly sensitive to the choice of $b$ and $k$, but is generally underpowered (relative to other choices) when $n/b=25$, $k/b=0.05$. This is because the test only has a small number of noisy TPDM estimates at its disposal. Conversely, the power tends to be marginally greatest when $n/b=50$ and $k/b=0.15$. However, with such a large effective sample size ($k_{\mathrm{total}}=\lfloor 0.15 \times 2500/50 \rfloor \times 50 = 350$) we might suspect that observations from the bulk are biasing the results. In this instance the bias appears to enhance the power, but it need not, since changes in dependence in the bulk and in the tail are generally two separate matters.

When dependence changes abruptly (SL-jump and HR-jump), the CM- and KS-based tests perform equally well. Upon further investigation, we find that for these models that paths $\{\hat{Z}_{ij}(t):t\in[0,1]$ are roughly $\wedge$-shaped curves attaining their suprema at $t=0.5$ i.e. when the changepoint occurs. Very loosely speaking, the CM-type test statistic corresponds to the largest area under these (squared) curves, while the KS-type test statistic corresponds to the largest supremum. By picturing $\{\hat{Z}_{ij}(t):t\in[0,1]$ as a triangle of width one and height $Z_{ij}(0.5)$, it becomes apparent that both test statistics are simply functions of $Z_{ij}(0.5)$ and thus contain equivalent information. For the linear dependence changes, the CM-based test is superior. 

Empirically, our test is more highly powered than @dreesStatisticalInferenceChanging2023. It achieves near full power for the SL-jump change; in the more challenging case of the HR-linear model, for which Drees' test is virtually powerless, our CM-type test discerns a signal more often that not. Initially, this might seem rather counterintuitive, since @dreesStatisticalInferenceChanging2023 leverages the full angular measure, whereas we rely solely on summary information. However, one can think of our method as imposing some
additional structure or information, namely that dependence is captured via the TPDM. When this assumption is fulfilled, a method that incorporates it will generally be superior to a fully non-parametric method that doesn't. In the case of the HR model this assumption does hold exactly, since dependence is fully characterised by the variogram $\Gamma(t)$, which is in one-to-one correspondence with the set of TPDMs. *(If more detail is needed to substantiate this claim, use Section 2.3 in https://arxiv.org/pdf/1207.6886 and Section 3 in Supp. Material of Cooley.)*

The Q-Q plots in @fig-samplesize-power show how the power improves as more data is acquired. For a given $\vartheta_1$, as $n$ increases, the curves shift further below the main diagonal. When the curves lies below the diagonal, this indicates that the rejection rate exceeds the nominal level and the test has power. 

```{r make-fig-bivariate-power}
#| label: fig-bivariate-power
#| fig-cap: "Empirical power against the dependence parameter $\\vartheta_1$. Based on 1000 simulations with $n=2,500$ and $d=2$."
#| fig-scap: "Empirical power against the dependence parameter $\\vartheta_1$."
#| fig-height: 6

p1 <- sim_results %>%
  filter(change_type %in% c("none", "jump"), d == 2, n == 2500, model == "log") %>%
  group_by(param1, n_blocks, k_frac, test_method, test_type) %>%
  summarise(power = 100 * mean(reject_H0)) %>%
  rename('n/b' = n_blocks, 'k/b' = k_frac) %>%
  ggplot(aes(x = param1, y = power, colour = test_method, shape = test_type, linetype = test_type)) +
  geom_point() +
  geom_line() +
  geom_hline(yintercept = 5, linetype = "dashed", colour = "lightgrey") +
  facet_grid(`k/b` ~ `n/b`, labeller = purrr::partial(label_both, sep = " = ")) +
  xlab(expression(vartheta[1])) +
  ylab("Empirical power (%)") +
  ggtitle("SL-jump") +
  scale_x_continuous(breaks = breaks_pretty(n = 3)) +
  scale_y_continuous(breaks = breaks_pretty(n = 5)) +
  scale_color_manual(labels = c("Drees", "Pawley"), values = c("red", "blue")) +
  scale_linetype_manual(labels = c("CM", "KS"), values = c(1, 3)) +
  scale_shape_manual(labels = c("CM", "KS"), values = 0:1) +
  theme_light() +
  labs(colour = "Test method", shape = "Test type", linetype = "Test type") +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        plot.title = element_text(hjust = 0.5, face = "bold", size = 10))

p2 <- sim_results %>%
  filter(change_type %in% c("none", "linear"), d == 2, n == 2500, model == "log") %>%
  group_by(param1, n_blocks, k_frac, test_method, test_type) %>%
  summarise(power = 100 * mean(reject_H0)) %>%
  rename('n/b' = n_blocks, 'k/b' = k_frac) %>%
  ggplot(aes(x = param1, y = power, colour = test_method, shape = test_type, linetype = test_type)) +
  geom_point() +
  geom_line() +
  geom_hline(yintercept = 5, linetype = "dashed", colour = "lightgrey") +
  facet_grid(`k/b` ~ `n/b`, labeller = purrr::partial(label_both, sep = " = ")) +
  xlab(expression(vartheta[1])) +
  ylab("Empirical power (%)") +
  ggtitle("SL-linear") +
  scale_x_continuous(breaks = breaks_pretty(n = 3)) +
  scale_y_continuous(breaks = breaks_pretty(n = 5)) +
  scale_color_manual(labels = c("Drees", "Pawley"), values = c("red", "blue")) +
  scale_linetype_manual(labels = c("CM", "KS"), values = c(1, 3)) +
  scale_shape_manual(labels = c("CM", "KS"), values = 0:1) +
  theme_light() +
  labs(colour = "Test method", shape = "Test type", linetype = "Test type") +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        plot.title = element_text(hjust = 0.5, face = "bold", size = 10))


p3 <- sim_results %>%
  filter(change_type %in% c("none", "jump"), d == 2, n == 2500, model == "hr") %>%
  group_by(param1, n_blocks, k_frac, test_method, test_type) %>%
  summarise(power = 100 * mean(reject_H0)) %>%
  rename('n/b' = n_blocks, 'k/b' = k_frac) %>%
  ggplot(aes(x = param1, y = power, colour = test_method, shape = test_type, linetype = test_type)) +
  geom_point() +
  geom_line() +
  geom_hline(yintercept = 5, linetype = "dashed", colour = "lightgrey") +
  facet_grid(`k/b` ~ `n/b`, labeller = purrr::partial(label_both, sep = " = ")) +
  xlab(expression(vartheta[1])) +
  ylab("Empirical power (%)") +
  ggtitle("HR-jump") +
  scale_x_continuous(breaks = breaks_pretty(n = 3)) +
  scale_y_continuous(breaks = breaks_pretty(n = 5)) +
  scale_color_manual(labels = c("Drees", "Pawley"), values = c("red", "blue")) +
  scale_linetype_manual(labels = c("CM", "KS"), values = c(1, 3)) +
  scale_shape_manual(labels = c("CM", "KS"), values = 0:1) +
  theme_light() +
  labs(colour = "Test method", shape = "Test type", linetype = "Test type") +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        plot.title = element_text(hjust = 0.5, face = "bold", size = 10))

p4 <- sim_results %>%
  filter(change_type %in% c("none", "linear"), d == 2, n == 2500, model == "hr") %>%
  group_by(param1, n_blocks, k_frac, test_method, test_type) %>%
  summarise(power = 100 * mean(reject_H0)) %>%
  rename('n/b' = n_blocks, 'k/b' = k_frac) %>%
  ggplot(aes(x = param1, y = power, colour = test_method, shape = test_type, linetype = test_type)) +
  geom_point() +
  geom_line() +
  geom_hline(yintercept = 5, linetype = "dashed", colour = "lightgrey") +
  facet_grid(`k/b` ~ `n/b`, labeller = purrr::partial(label_both, sep = " = ")) +
  xlab(expression(vartheta[1])) +
  ylab("Empirical power (%)") +
  ggtitle("HR-linear") +
  scale_x_continuous(breaks = breaks_pretty(n = 3)) +
  scale_y_continuous(breaks = breaks_pretty(n = 5)) +
  scale_color_manual(labels = c("Drees", "Pawley"), values = c("red", "blue")) +
  scale_linetype_manual(labels = c("CM", "KS"), values = c(1, 3)) +
  scale_shape_manual(labels = c("CM", "KS"), values = 0:1) +
  theme_light() +
  labs(colour = "Test method", shape = "Test type", linetype = "Test type") +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        plot.title = element_text(hjust = 0.5, face = "bold", size = 10))

ggarrange(p1, p2, p3, p4, ncol = 2, nrow = 2, common.legend = TRUE, legend = "right")
```

```{r}
#| label: fig-samplesize-power
#| fig-cap: "QQ-plots for the KS p-values with varying sample size. Based on 1000 simulations from the HR-jump (left) and HR-linear (right) models with $\\vartheta_1\\in\\{0.5,2\\}$, $n/b=25$ and $k/b=0.1$."
#| fig-scap: "Empirical power against the sample size $n$."
#| fig-height: 2.5

sim_results %>%
  filter(change_type != "none", n < 10^6, d == 2, model == "hr", test_method == "pawley", test_type == "ks", n_blocks == 25, k_frac == 0.1) %>%
  ggplot(aes(sample = 1 - CPAT:::pkolmogorov(test_stat)^(d*(d-1)/2), colour = as.factor(n), linetype = as.factor(param1))) +
  stat_qq(distribution = qunif, geom = "path") +
  geom_abline(intercept = 0, slope = 1, linetype = 1, colour = "darkgrey") +
  facet_grid(~ change_type, labeller = labeller(.default = str_to_title)) +
  xlab("Uniform quantiles") +
  ylab("Sample quantiles") +
  labs(colour = expression(n), linetype = expression(vartheta[1])) +
  scale_x_continuous(limits = c(0, 1), expand = c(0, 0), breaks = breaks_pretty(n = 1)) +
  scale_y_continuous(limits = c(0, 1), expand = c(0, 0.005), breaks = breaks_pretty(n = 3)) +
  scale_color_manual(labels = c("2,500", "5,000", "10,000"), values = c("red", "blue", "darkgreen")) +
  scale_linetype_manual(labels = c("0.5", "2"), values = c(1, 2)) +
  theme_light() +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        plot.title = element_text(hjust = 0.5, face = "bold", size = 10))
```

*Discussion of computation time here, referring to @fig-computation-time.*

```{r make-fig-computation-time}
#| label: fig-computation-time
#| fig-cap: "Average computation time across numerical experiments."
#| fig-scap: "Average computation time across numerical experiments."
#| fig-height: 5
#| 
p1 <- sim_results %>%
  filter(model == "log", d == 2, test_type == "ks", n < 10^6) %>%
  ggplot(aes(x = elapsed_time, y = as.factor(n), fill = as.factor(k_frac))) +
  geom_boxplot(outliers = FALSE) + 
  facet_grid(as.factor(n_blocks) ~ test_method, scales = "free", labeller = labeller(.default = str_to_title)) + 
  xlab("Computation time (s)") +
  ylab("Sample size, n") +
  coord_flip() + 
  labs(fill = expression(k/b)) + 
  scale_fill_discrete_sequential("Red-Blue") +
  theme_light()

p2 <- sim_results %>%
  filter(test_method == "pawley", test_type == "ks", n < 10^6) %>%
  mutate(k_total = k * n_blocks) %>%
  group_by(d, k_total) %>%
  summarise(median_time = median(elapsed_time),
            lower_time = quantile(elapsed_time, probs = 0.25),
            upper_time = quantile(elapsed_time, probs = 0.75)) %>%
  ungroup() %>%
  filter(k_total != 375) %>%
  ggplot(aes(x = k_total, colour = as.factor(d))) +
  geom_point(aes(y = median_time)) + 
  geom_path(aes(y = median_time)) +
  scale_x_continuous(breaks = breaks_pretty(n = 8)) +
  scale_y_continuous(trans = "log10") +
  geom_ribbon(aes(ymin = lower_time, ymax = upper_time, fill = as.factor(d)), alpha = 0.15, colour = NA) +
  xlab(expression("Effective sample size," ~ k[total] == nk/b)) +
  ylab("Computation time (s)") +
  labs(colour = expression(d), fill = expression(d)) + 
  scale_color_manual(labels = c("2", "5", "10", "25"), values = c("red", "blue", "darkgreen", "orange")) +
  scale_fill_manual(labels = c("2", "5", "10", "25"), values = c("red", "blue", "darkgreen", "orange")) +
  theme_light()
  
ggarrange(p1, p2, ncol = 1, nrow = 2, common.legend = FALSE)
```

## No free lunch: constant TPDM with changing dependence {#sec-constant-tpdm}

Our proposed extension to @dreesStatisticalInferenceChanging2023 affords many advantages, most notably the ability to conduct tests in high dimensions. The price paid is that we forgo the ability to detect TPDM-invariant dependence changes. (The existence of such changes is consequence of the many-to-one correspondence between angular measures and TPDMs.) For this class of alternatives our test will be inherently predisposed to commit Type II errors. In this section, we illustrate this flaw by constructing a sub-class of examples based on a time-dependent version of the max-linear model. 

Suppose $\{\bm{X}(t):t\in[0,1]\}$ is a $d$-dimensional stochastic process defined by
\begin{equation}\label{eq-time-dependent-max-linear}
  \bm{X}(t) = A(t) \times_{\max}\bm{Z}(t), \qquad A(t)=A_0\ind\{t < 0.5\} + A_1\ind\{t \geq 0.5\}.
\end{equation}
The stochastic innovations process $\{\bm{Z}(t)=(Z_1(t),\ldots,Z_q(t)):t\in[0,1]\}$ is a collection of independent random vectors; for any $t\in[0,1]$ the $q\geq 1$ components of $\bm{Z}(t)$ are independently Fréchet distributed with shape parameter equal to 2. The dependence structure of $\bm{X}(t)$ is characterised by the parameter matrix $A(t) = (a_{ij}(t))\in \R_+^{d\times q}$. Under the model \eqref{eq-time-dependent-max-linear}, the dependence parameter undergoes a jump-change from $A_0\in \R_+^{d\times q}$ to $A_1\in \R_+^{d\times q}$ at time $t=0.5$. More flexible models can easily be conceived, whereby $A(t)$ evolves smoothly, perhaps even with a varying number of factors $q=q(t)$, but the simple model above will suffice for our aims. The local angular measure associated with \eqref{eq-time-dependent-max-linear} can be expressed in terms of the columns $\bm{a}_1(t),\ldots,\bm{a}_q(t)\in\R_+^d$ of $A(t)$ as
\begin{equation*}
H(\cdot\,;t) = \sum_{j=1}^q \|\bm{a}_j(t)\|_2^2 \delta_{\bm{a}_j(t)/\|\bm{a}_j(t)\|_2}(\cdot).
\end{equation*}
The local TPDM is given by $\Sigma(t) = A(t)A(t)^T$ and the diagonal and off-diagonal entries of its asymptotic covariance $V(t)$ matrix are given by @krali
\begin{equation}\label{eq-time-dependent-max-linear-V}
k\mathrm{Cov}(\hat{\sigma}_{ij}(t),\hat{\sigma}_{lm}(t)) \to 
\begin{cases}
  d\sum_{s=1}^q \frac{a_{is}(t)^2 a_{js}(t)^2}{\|\bm{a}_s(t)\|_2^2} - \sigma_{ij}(t)^2, & i=l, j=m, \\
  d\sum_{s=1}^q \frac{2a_{is}(t)a_{js}(t)a_{ls}(t)a_{ms}(t)}{\|\bm{a}_s(t)\|_2^2} - 2\sigma_{ij}(t)\sigma_{lm}(t) , & \text{otherwise}.
\end{cases}
\end{equation}

If $A_0$ and $A_1$ are distinct (up to permutations of their columns) yet carefully chosen so that $A_0A_0^T=A_1A_1^T$ and \eqref{eq-time-dependent-max-linear-V} yields identical asymptotic covariances, then the alternative hypothesis \eqref{eq-changing-ext-dep-alt-hyp} is true but the convergences \eqref{eq-asymptotic-test-stat} still hold. Finding non-trivial (i.e. $q>2$) pairs $A_0,A_1$ by hand would be extremely laborious, if not impossible, so we resort to a computational approach. We generate $N\gg 1$ candidate $d\times q$ matrices with uniformly distributed entries; the rows of each matrix are subsequently normalised to ensure the resulting TPDM is properly scaled. Then we search for pairs of matrices satisfying (within some small tolerance) the required conditions. Using this procedure with $d=2$, $q=20$, and $N=50,000$, we find a suitable matrix pair for which $\sigma_{12}(t)=0.1000$ and $k\mathrm{Var}(\hat{\sigma}_{12}(t)) \to 0.060$, to three decimal places. We generate 1000 datasets, each with $n=10,000$ samples, from the model \eqref{eq-time-dependent-max-linear} using the\texttt{SpatialExtremes} package. For each dataset, we apply our test and that of @dreesStatisticalInferenceChanging2023 with $b=400$ and $k=40$. 

The diagnostic plots in @fig-counterexample-maxlinear-pawley illustrate the computations underlying our testing procedure when applied to one of these datasets. The top-left panel depicts the empirical local TPDM over time. (Since $d=2$, there is only one component pair to consider.) The range of values of $\hat{\sigma}_{12}(t)$ is consistent with \eqref{eq-emp-local-tpdm-asymptotic}, which implies that 
\begin{equation*}
\left(0.1000 - \Phi^{-1}(0.975)\sqrt{\frac{0.060}{40}}, 0.1000 + \Phi^{-1}(0.975)\sqrt{\frac{0.060}{40}}\right) \approx (0.724, 0.876)
\end{equation*}
represents a 95\% asymptotic confidence interval for $\sigma_{12}(t)$. Indeed, the empirical coverage of the interval, based on the $1000 \times n/b = 25000$ estimates of $\sigma_{12}(t)$ from across the full set of simulations, equals 93.56\%. There is no temporal trend in the blocks' TPDMs, so the integrated TPDM (top-right panel) is a straight line and the test process $Z_{12}(t)$ (bottom-left) resembles a typical Brownian bridge sample path. The bottom-right panel depicts $\int_0^t |\hat{Z}_{12}(s)|\,\dee s$ (CM, upper sub-panel) and $\sup_{0\leq s \leq t} |\hat{Z}_{12}(s)|$ (KS, lower sub-panel) as functions of $t$. The maximal values of these processes do not exceed the associated critical values at the 5\% level, marked by the dashed lines. We conclude there is insufficient evidence to conclude that dependence is changing and commit a Type II error. The empirical Type II error rates across all 1000 replications of the experiment are 94.5\% (CM) and 96.5\% (KS). As expected, the empirical power of the test is approximately the desired Type I error rate.

@fig-counterexample-maxlinear-drees shows the analogous plots corresponding to Drees' method applied to the same data. The left-hand plot depicts the empirical integrated angular measure as a function of $t$. Each curve corresponds to a particular set $A_y\in\mathcal{A}$, with darker colours indicating larger values of $y$. Close inspection of these curves reveals a slight kink at $t=0.5$. The dependence change is more apparent in the middle panel, which depicts the corresponding $|\mathcal{A}|$-dimensional test process. This process is analogous to \eqref{eq-test-process}, but its interpretation is less straightforward because the curves are cross-correlated. Computing the relevant time-integrals of these processes yields the curves in the right-hand plot. For both the CM- and KS-based tests, there exists a curve that enters the rejection region demarcated by the dashed lines, so according to either test we would (correctly) reject the null hypothesis at the 5\% level. Upon repeating this 1000 times, the test's empirical power is found to be 100\% (CM) and 99.8\% (KS).

```{r make-fig-counterexample-pawley}
#| label: fig-counterexample-maxlinear-pawley
#| fig-cap: "Diagnostic plots for our test for data from \\eqref{eq-time-dependent-max-linear} with $n=10,000$, $b=400$, $k=40$."
#| fig-height: 3.5

readRDS(file = file.path("scripts", "changing-ext-dep", "results", "constant-tpdm-pawley-example.RDS")) %>%
  extract2("data") %>%
  plot_test_pawley(variable_scheme = "facet")
```


```{r make-fig-counterexample-maxlinear-drees}
#| label: fig-counterexample-maxlinear-drees
#| fig-cap: "Diagnostic plots for Drees' test for data from \\eqref{eq-time-dependent-max-linear} with $n=10,000$, $b=400$, $k=40$. Each curve represents a set $A_y$ with darker colours indicating larger values of $y$."
#| fig-height: 2

readRDS(file = file.path("scripts", "changing-ext-dep", "results", "constant-tpdm-drees-example.RDS")) %>%
  extract2("data") %>%
  plot_test_drees()
```

## Application: extreme Red Sea surface temperatures

We now apply our methodology to test for changing dependence in extreme Red Sea surface temperature anomalies. The dataset has been widely studied in the extremes community, primarily because it was the focus of the EVA 2019 Data Challenge but also because extreme temperatures are related to ecological issues such as coral bleaching. Further details about the data collection and pre-processing can be found in Huser (2020).

Previous investigations by Simpson and Wadsworth (2020) and Huser (2020) conclude that surface temperature extremes exhibit differing behaviour in the north and south, so it is advisable to treat these areas separately. We divide the spatial domain into northerly and southerly sub-regions, each comprising 70 sites whose are shown in @fig-redsea-sites. 

```{r}
#| label: fig-redsea-sites
#| fig-cap: "Locations of the 70 sites in each of the two sub-regions in the Red Sea."
#| fig-height: 3.2

redsea <- load_red_sea_temp(alpha = 2)

ggplot() +
  geom_polygon(data = map_data("world"),
               aes(x = long, y = lat, group = group),
               color = "white", fill = "#d8c596") +
  coord_fixed(0.6, xlim = c(32, 44), ylim = c(12, 30)) +
  xlab("Longitude") +
  ylab("Latitude") +
  scale_x_continuous(breaks = breaks_pretty(n = 3)) +
  scale_y_continuous(breaks = breaks_pretty(n = 5)) +
  geom_point(data = redsea$coord, aes(x = longitude, y = latitude, colour = region),
             shape = 16) +
  scale_color_manual(labels = c("None", "North", "South"), values = c(alpha("grey", 0.6), "red", "darkblue")) +
  labs(colour = "Region") +
  theme_light() +
  theme(panel.background = element_rect(fill = alpha("skyblue", 0.7)),
        legend.position = "none",
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())
```


At any particular location, daily maxima are known to occur in (temporal) clusters, meaning high temperatures may persist across several days (Simpson and Wadsworth, 2020). We address this by working with weekly maxima, so that observations are approximately independent over time. This yields $n=1605$ samples spanning approximately 31 years. Let $X_i^{(\text{north})}(t)$ and $X_i^{(\text{south})}(t)$ denote the surface temperature anomaly (on stationary Fréchet margins) at site $i\in\{1,\ldots,70\}$ and time $t\in[0,1]$ in the north and south sub-regions, respectively. 

Our goal is to determine whether either of
\begin{equation*}
\bm{X}^{(\text{north})}(t)=\{X_i^{(\text{north})}(t):i=1,\ldots,70\}, \qquad 
\bm{X}^{(\text{south})}(t)=\{X_i^{(\text{south})}(t):i=1,\ldots,70\}
\end{equation*}
exhibit evidence for stationary or changing extremal dependence. To this end, we will apply our test using $b=107$ and $k=20$, yielding 15 blocks and an effective sample size of $k_{\mathrm{total}}=15\times 20=300$. The rank condition (and other considerations) restricts us to testing up to 17 sites at a time; it is not possible/advisable to test for changing dependence in each region using all 70 sites. Our strategy will be to repeatedly sample $2 \leq d \leq 17$ sites from each region. We apply this procedure $N=1000$ times for $d\in\{5, 10, 15\}$, perform our test, and collate the resulting p-values. Their distributions are shown in @fig-redsea-p-values. 

Rough summary of conclusions: North shows evidence of changing dependence, South not so much; results for $d=15$ are unreliable as convergence unlikely (based on earlier tables etc.); CM has higher rejection rate than KS (aligns with sim studies that shows CM has greater power than KS when dependence change is gradual, as is likely the case here). 
For $d=5$, the p-values are strongly skewed towards zero, resulting in a rejection rate of approximately 60\% (for both KS and CM). 

```{r}
#| label: fig-redsea-p-values
#| fig-cap: "Blah."
#| fig-height: 4

list(file.path("scripts", "changing-ext-dep", "results", "redsea-north-p-values.RDS"),
     file.path("scripts", "changing-ext-dep", "results", "redsea-south-p-values.RDS")) %>%
    lapply(readRDS) %>%
    bind_rows() %>%
    mutate(d = as.factor(paste("d", d, sep = " = "))) %>%
    ggplot(aes(x = p_value)) +
    geom_histogram(breaks = seq(0, 1, 0.025), color = "black", fill = "grey", linewidth = 0.3) +
    geom_vline(xintercept = 0.05, colour = "darkblue", linetype = "dashed", linewidth = 0.3) +
    facet_nested(test_type ~  region + d, scales = "free", labeller = labeller(region = str_to_title, .rows = toupper), nest_line = element_line(colour = "white")) +
    scale_x_continuous(expand = c(0, 0), breaks = breaks_pretty(n = 1)) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
    xlab("p-value") +
    ylab("Frequency density") +
    theme_light() +
    theme(panel.grid.minor = element_blank(),
          panel.grid.major = element_blank())
```


```{r}
#| label: fig-redsea-test-example
#| fig-cap: "Diagnostic plots for our test, based $b=107$ and $k=20$, applied to data from $d=10$ randomly selected northerly sites in the Red Sea. Each curve corresponds one of the $\\mathcal{D}=45$ component pairs."
#| fig-height: 4

set.seed(1)

north_sites <- redsea$coord %>%
  mutate(loc_index = row_number()) %>%
  filter(region == "north") %>%
  slice_sample(n = 10) %>%
  pull(loc_index)
    
redsea$X[1:1605, north_sites] %>%
  as_tibble() %>%
  test_pawley(b = 107, k = 20, return_all = TRUE) %>%
  extract2("data") %>%
  plot_test_pawley(variable_scheme = "colour")
```

## Extensions and modifications

### Alternative dependence measures

Our method considers the time-evolution of the dependence between $X_i$ and $X_j$ according to the measure
\begin{equation}\label{eq-tdpm-general-f}
\sigma_{ij}(t)=\lim_{r\to\infty}\mathbb{E}[f(\bm{\Theta}(t))\mid R(t) > r],
\end{equation}
where $f:\mathbb{S}_+^{d-1}\to\R_+$ is defined by $f(\bm{\Theta})=d\theta_i\theta_j$. However, the EDM/TPDM is just one measure of extremal dependence among a large class. Alternative measures can be generated by replacing $f$ in \eqref{eq-tdpm-general-f} with other functions $g:\mathbb{S}_+^{d-1}\to\R_+$ [@larssonExtremalDependenceMeasure2012]. Provided $g$ satisfies the conditions of Theorem 4 in @kluppelbergEstimatingExtremeBayesian2021, the theory underpinning our testing methodology holds. That these alternative measures lack the nice properties of the TPDM, such as positive definiteness, is not particularly relevant for the task-at-hand. The circumstances under which a particular measure is inferior/superior (in terms of power, say) to others is governed by the nature of the associated function $g$. A practitioner working in a particular setting, where the dependence structures and dependence changes tend to be of a certain nature, might wish to tailor the dependence measure to suit their purposes. This could be achieved by running a series of numerical experiments, designed to mimic the scenarios they typically encounter, and choosing $g$ optimally among some (parametric) subfamily according to some performance metric.

*I could illustrate this process with a simple toy example, e.g. take $g(\bm{\theta};\gamma)=d\theta_i^\gamma\theta_j^\gamma$ and find $\gamma\in(0,4]$ that achieves maximal empirical power on a particular model.*

### Changepoint detection

Our primary objective was to devise a test to ascertain whether or not an assumption of constant tail dependence is reasonable. In certain applications (e.g. finance), it may be more interesting to ask *when*, not if, dependence has changed. This is the realm of changepoint detection. Suppose the angular measure of $\{\bm{X}(t):t\in[0,1]\}$ is given by $H(t)=H_0\ind\{t\leq \tau\} + H_1\ind\{t >\tau\}$ for some $\tau\in(0,1)$. Then 
\begin{equation*}
\hat{\tau} = \ldots
\end{equation*}
is a CUSUM-type estimator of $\tau$. *Discuss (and illustrate) how this estimator is biased towards the centre of the time interval, and the modifications that would be needed to remedy this.*

### Robustness

*Test robustness to serial dependence. (e.g. simulation from AR process)*

## Other things could do

*Do example where dependence only changes in a subset of components. How does power vary against proportion of pairs that undergo change?*
