# Compositional perspectives on extremes

```{r compositional-load-packages}
#| include: false
library(tidyverse)
library(tidyr)
library(magrittr)
library(scales)
library(ggh4x)
library(ggpubr)
library(colorspace)
library(pbapply)
library(kableExtra)
library(reshape2)
library(compositions)
library(Ternary)
library(SpatialExtremes)
library(caret)
library(lattice)

options(dplyr.summarise.inform = FALSE)
options(knitr.kable.NA = "")
```

```{r compositional-source-functions}
#| include: false
sapply(list.files(path = "R/compositional", pattern = "*.R", full.names = TRUE, recursive = TRUE), source)
sapply(list.files(path = "R/general", pattern = "*.R", full.names = TRUE, recursive = TRUE), source)
```

The primary object of interest in multivariate extremes is the angular measure, which represents the limiting conditional distribution of $\bm{\Theta}\mid R>t$, where $\bm{\Theta}:=\bm{X}/\|\bm{X}\|_\alpha$ and $R:=\|\bm{X}\|_\alpha$, as $t\to\infty$. Taking $\alpha=1$ and $\bm{X}\in\mathrm{RV}_+^d(1)$ on unit Fréchet margins, the following statements are true:

1. $\bm{\Theta}$ lies in the $(d-1)$-dimensional simplex $\mathbb{S}_+^{d-1}=\{\bm{\theta}\in\R^d : \theta_j\geq 0, \sum_{j=1}^d \theta_j=1\}$ in $\R^d$. 
2. In the limit as $t\to\infty$, the angle of threshold exceedances $\bm{\Theta}\mid R>t$ is independent of $R$. 
3. The (normalised) angular measure satisfies $\int_{\mathbb{S}_+^{d-1}}\theta_j\,\dee H(\bm{\theta})=1/d$ for all $j=1,\ldots,d$.

Property (i) states that $\bm{\Theta}$ is a $d$-part *random composition* with non-negative components summing to unity. In his seminal paper, @aitchisonStatisticalAnalysisCompositional1982 contended that analysing such data using standard methodology designed for unconstrained vectors can lead to misleading inferences. Compositional data analysis (CoDA) subsequently emerged as a discipline for developing statistical theory and techniques tailored to account for the geometry of the simplex. Statement (ii) expresses that, in the limit, MRV random vectors satisfy a fundamental principle of CoDA called *scale invariance*: the distribution of the composition (angular component) of is independent of the absolute size (radial component). Finally, the moment constraint (iii) means that the centre of mass of any valid angular measure (with respect to $\|\cdot\|_1$) lies at the barycentre of the simplex.

There is a clear connection between compositional data analysis and multivariate extreme value statistics. We are not the first to notice this link. @colesModellingExtremeMultivariate1991 remark that parametric CoDA models would be useful for modelling extremal dependence, were it not for the fact that they typically violate the moment constraint. In a paper outlining potential future avenues for research in extremes, @longin2016 suggests applying "PCA for compositional data \textellipsis to the pseudo-angles" to "disentangle dependence into components \ldots of practical interest". Finally, @serranoSemiparametricBivariateExtremevalue2022 leverage CoDA techniques (e.g. log-ratio transformations and compositional splines) to construct bivariate extreme value copulas. 

This chapter aims to explore the link between these two statistical disciplines more thoroughly. In particular, we apply a CoDA lens to two statistical learning problems within multivariate extremes: tail dimension reduction via principal components analysis [@clemenconRegularVariationHilbert2024; @cooleyDecompositionsDependenceHighdimensional2019; @dreesPrincipalComponentAnalysis2021] and binary classification in extreme regions [@jalalzaiBinaryClassificationExtreme2018]. We demonstrate that off-the-shelf CoDA methods are readily applicable to these problems and compare their performance against existing state-of-the-art methods from the extremes literature.

## Compositional data analysis

@aitchisonStatisticalAnalysisCompositional1982 argues that standard multivariate data analysis techniques are inappropriate for modelling compositional data because they are designed for unconstrained data. Neglecting the compositional constraint causes an array of difficulties: spurious correlations [@pearsonMathematicalContributionsTheory1897; @aitchisonStatisticalAnalysisCompositional1982], a failure to capture the marked curvature characteristic of compositional data sets [@aitchisonPrincipalComponentAnalysis1983], or contradictory conclusions between analyses depending on which components are included in the composition [@aitchisonStatisticalAnalysisCompositional1986]. These issues are addressed by devising a statistical framework tailored to the algebraic-geometric structure of the underlying sample space: the unit simplex. This section reviews the basic concepts and principles of CoDA. 

### Compositions

Compositional analysis concerns vectors with strictly positive components for which all relevant information is *relative*, i.e. conveyed by ratios between components.

:::{#def-composition}
Vectors $\bm{x},\bm{y}\in\R_+^d:=(0,\infty)^d$ are *compositionally equivalent*, denoted $\bm{x}\sim\bm{y}$, if there exists $c>0$ such that $\bm{y}=c\bm{x}$.
:::

The equivalence relation $\sim$ defines equivalence classes on $\R_+^d$. 

:::{#def-closed-composition}
For $\bm{x}\in\R_+^d$, the compositional class $[\bm{x}]:=\{c\bm{x}:c>0\}$ is represented on the $d$-part unitary simplex by its *closed composition* given by
\begin{equation*}
    \mathcal{C}\bm{x} := \frac{(x_1,\ldots,x_d)}{\sum_{i=1}^d x_i}.
\end{equation*}
:::

### Aitchison geometry

Adhering to the principles propounded by @aitchisonStatisticalAnalysisCompositional1986 necessitates introducing alternative notions of mean/variance, distance, projections, etc. that make sense for compositions. Formally, this involves constructing a Hilbert space structure on $\mathbb{S}_+^{d-1}$ [@aitchisonStatisticalAnalysisCompositional1986; @pawlowsky-glahnGeometricApproachStatistical2001]. This is achieved by defining a vector space structure on $\mathbb{S}_+^{d-1}$ and endowing it with a suitable inner product, norm and distance.

:::{#def-aitchison-vector-space}
Let $\bm{x},\bm{y}\in\mathbb{S}_+^{d-1}$ be closed compositions and $\alpha\in\R$ a scalar. The perturbation and power operations are defined by
\begin{equation*}
    \bm{x} \oplus \bm{y} = \mathcal{C}(x_1y_1,\ldots,x_dy_d),\qquad \alpha \odot \bm{x} = \mathcal{C}(x_1^\alpha,\ldots,x_d^\alpha).
\end{equation*}
:::

It is straightforward to show that $(\mathbb{S}_+^{d-1},\odot,\oplus$) is a real vector space. The additive identity and inverse elements are $\bm{e}_a:=\mathcal{C}(1,\ldots,1)$ and $\ominus \bm{x} := \mathcal{C}(x_1^{-1},\ldots,x_d^{-1})$, respectively.

:::{#def-clr}
The \textit{centred log-ratio (CLR) transformation} is
\begin{equation*}
    \mathrm{clr}:\R^d\to \R^d,\qquad \bm{x}\mapsto \log\left(\frac{\bm{x}}{\bar{g}(\bm{x})}\right),
\end{equation*}
where $\bar{g}(\bm{x}):=(\prod_{i=1}^d x_i)^{1/d}$ denotes the geometric mean of the components of $\bm{x}$. 
:::

CLR-transformed vectors lie in the hyperplane $\mathcal{T}^{d-1}:=\{\bm{y}\in\R^d : y_1+\ldots+y_d=0 \}\subset\R^d$. The transformation can be inverted to recover the original (closed) composition via
\begin{equation*}
    \mathrm{clr}^{-1} : \mathcal{T}^{d-1} \to \mathbb{S}_+^{d-1}, \qquad \bm{v} \mapsto \mathcal{C}\exp(\bm{v}).
\end{equation*}

:::{#def-aitchison-metric-space}
Let $\bm{x},\bm{y}\in\mathbb{S}_+^{d-1}$ be closed compositions. Let $\langle \cdot,\cdot\rangle_e$ denote the Euclidean inner product in $\R^d$. The *Aitchison inner product* in $\mathbb{S}_+^{d-1}$ is
\begin{equation*}
      \langle \bm{x},\bm{y}\rangle_a := \langle \mathrm{clr}(\bm{x}),\mathrm{clr}(\bm{y})\rangle_e = \sum_{i=1}^d \log\left(\frac{x_i}{\bar{g}(\bm{x})}\right) \log\left(\frac{y_i}{\bar{g}(\bm{y})}\right).
\end{equation*}
The *Aitchison norm* and *Aitchison distance* are the metric elements induced by $\langle \cdot,\cdot \rangle_a$, that is
\begin{equation}\label{eq-aitchison-metric}
      \|\bm{x}\|_a := \langle \bm{x},\bm{x} \rangle_a^{1/2}, \qquad d_a(\bm{x},\bm{y}) := \|\bm{x} \ominus \bm{y}\|_a.
\end{equation} 
:::

The CLR-transformation is an isometry between $\mathcal{T}^{d-1}$ equipped with Euclidean geometry and $\mathbb{S}_+^{d-1}$ equipped with Aitchison geometry.

:::{#def-composition-centre-variance}
The centre and total variance of a random composition $\bm{X}$ are given by 
\begin{align*}
    \mathrm{cen}_a(\bm{X}) &:= \argmin_{y\in\mathbb{S}_+^{d-1}}\mathbb{E}[d_a^2(\bm{X},\bm{y})] = \mathcal{C}(\exp(\mathbb{E}[\log(\bm{X})]), \\
        \mathrm{totVar}_a(\bm{X}) &:= \mathbb{E}[d_a^2(\bm{X},\mathrm{cen}_a(\bm{X}))] = \sum_{j=1}^d \mathrm{Var}([\mathrm{clr}(\bm{X})]_j).
\end{align*}
:::

### Compositional principal components analysis

Compositional principal component analysis (CoDA-PCA) aims at finding low-dimensional descriptions of compositional data which retain most of the variability in the original data [@aitchisonPrincipalComponentAnalysis1983]. Classical PCA based on Euclidean geometry is ill-suited to this task for two main reasons. First, PCA is typically used as an exploratory tool for understanding the correlation structure among a set of variables, but the compositional constraint places restrictions on this structure and spurious correlations arise when these are not properly accounted for [@aitchisonStatisticalAnalysisCompositional1982]. Second, the Hilbert space in which PCA is conducted should conform to the underlying sample space to facilitate a consistent and interpretable analysis. For example, compositional data often exhibit curvature that cannot be captured by Euclidean straight lines, and Euclidean projections of the data may lie outside of the simplex [@aitchisonPrincipalComponentAnalysis1983].

Let $\bm{X}=(X_1,\ldots,X_d)$ denote a $d$-part centred random composition with finite second order moments, that is $\mathrm{cen}_a(\bm{X})=\bm{e}_a$ and $\mathbb{E}[\|\bm{X}\|_a^2]<\infty$. Let $\Gamma=\mathrm{Cov}(\mathrm{clr}(\bm{X}))$ denote the CLR-covariance matrix and $\Gamma=U\Lambda U^T$ its eigendecomposition, where $\Lambda$ is a diagonal matrix of eigenvalues $\lambda_1 \geq \lambda_2 \geq \ldots \geq \lambda_{d-1} \geq \lambda_d = 0$ and $U$ is an orthonormal $d\times d$ matrix whose columns are the eigenvectors $\bm{u}_1,\ldots,\bm{u}_{d-1}\in\mathcal{T}^{d-1}$ and $\bm{u}_d\propto\bm{1}_d$. CoDA-PCA consists in retaining the leading $p\leq d-1$ (back-transformed) eigenvectors to account for a desired proportion of the total variability of $\bm{X}$ [@wangPrincipalComponentAnalysis2015]. For any linear subspace $V\subset\mathbb{S}_+^{d-1}$, let $\Pi_V$ denote the orthogonal projection onto $V$. Then,
\begin{equation*}
    \mathcal{V}_p := \mathrm{span}_a(\mathrm{clr}^{-1}(\bm{u}_1),\ldots,\mathrm{clr}^{-1}(\bm{u}_p)) := \left\lbrace \bigoplus_{j=1}^p (\alpha_j \odot \mathrm{clr}^{-1}(\bm{u}_j)) : \alpha_1,\ldots,\alpha_p \in\R \right\rbrace
\end{equation*}
minimises the expected squared (Aitchison) reconstruction error among all $p$-dimensional linear simplicial subspaces. The low-dimensional approximation $\Pi_{\mathcal{V}_p}\bm{X}$ accounts for a proportion
\begin{equation*}
    \frac{\mathrm{totVar}(\Pi_{\mathcal{V}_p}\bm{X})}{\mathrm{totVar}(\bm{X})} = \frac{\sum_{j=1}^p \lambda_j}{\sum_{j=1}^{d-1}\lambda_j}
\end{equation*}
of the total variance. The compositional line $\{\tau \odot \mathrm{clr}^{-1}(\bm{v}_j) : \tau\in\R\}\subset\mathbb{S}_+^{d-1}$ represents the trend described by the $j$th principal component.

### Compositional classification based on the $\alpha$-metric

The $k$-nearest neighbours ($k$-NN) algorithm is a simple, popular non-parametric classifier [@hastieElementsStatisticalLearning2009]. Suppose we observe training samples $(\bm{x}_i,y_i),\ldots,(\bm{x}_n,y_n)$ of $(\bm{X},Y)$, where $Y\in\{0,1\}$ is the binary class label of $\bm{X}$. In the classification step, a new test observation $\bm{x}^\star$ is allocated to the majority class among its $k$ nearest neighbours, with ties broken randomly or according to some other pre-specified criterion. The tuning parameter $k\geq 1$ determines the flexibility of the classification boundaries and is usually selected by a cross-validation procedure.

The notion of neighbours implicitly assumes an underlying metric, which is typically taken to be the Euclidean distance. However, if $\bm{X}$ is a compositional random vector, then the CoDA philosophy dictates that a simplicial distance measure should be preferred. The Aitchison metric \eqref{eq-aitchison-metric} is an obvious choice. However, @greenacreChiPowerTransformationValid2024 argues that in the supervised setting, where an objective performance criterion (e.g. the out-of-sample classification error rate) is available, we should not be wedded to this choice. In this spirit, @tsagrisImprovedClassificationCompositional2016 propose a compositional classification algorithm called $\alpha$-transformed compositional $k$-nearest neighbours, henceforth denoted $k$-NN($\alpha$). The additional tuning parameter $\alpha\in\R$ relates to a Box-Cox-type data transformation upon which their proposed simplicial distance is based [@tsagrisDatabasedPowerTransformation2011].

:::{#def-alpha-transformation}
For $\alpha \neq 0$, the $\alpha$-transformation of a composition $\bm{x}\in\mathbb{S}_+^{d-1}$ is 
\begin{equation*}
        \bm{z}_\alpha : \mathbb{S}_+^{d-1} \to \R^d, \qquad \bm{x} \mapsto H \cdot \left(\frac{d (\alpha\odot \bm{x}) - \bm{1}_d}{\alpha}\right),
\end{equation*}
where $H$ is any $(d-1)\times d$ real matrix with orthonormal rows. For $\alpha=0$, we define $\bm{z}_0(\bm{x}):=\lim_{\alpha\downarrow 0} \bm{z}_\alpha(\bm{x})$. 
:::

Typically $H$ is chosen as the Helmert matrix with its first row removed, but in any case $k$-NN($\alpha$) is invariant to this choice. The $\alpha$-transformation induces a metric on $\mathbb{S}_+^{d-1}$ in a similar fashion to the CLR-transformation.

:::{#def-alpha-metric}
Let $\bm{x},\bm{y}\in\mathbb{S}_+^{d-1}$ be closed compositions. For $\alpha\in\R$, the $\alpha$-metric is defined as
\begin{equation*}
  d_\alpha(\bm{x},\bm{y}) := \|\bm{z}_\alpha(\bm{x}) - \bm{z}_\alpha(\bm{y})\|_e
\end{equation*}
:::

The special cases $\alpha=0$ and $\alpha=1$ correspond to the Aitchison and Euclidean distances (up to a scalar multiple), respectively, that is $d_0(\bm{x},\bm{y})= d_a(\bm{x},\bm{y})$ and $d_1(\bm{x},\bm{y})= d \cdot d_e(\bm{x},\bm{y})$. This means that the family of $k$-NN($\alpha$) classifiers encompasses Euclidean- and Aitchison-based $k$-NN classifiers, but some alternative value of $\alpha$ may be selected if it achieves superior performance. One can easily devise analogues of other classifiers, such as $\alpha$-transformed support vector machines ($\alpha$-SVM) and $\alpha$-transformed random forests ($\alpha$-RF) [@tsagrisImprovedClassificationCompositional2016]. 

## Compositional PCA for extremes

### Framework and motivation

Suppose $\bm{X}=(X_1,\ldots,X_d)\in\mathrm{RV}_+^d$ is an $\R_+^d$-valued random vector with tail index $\alpha$. Let $H_{\bm{X}}$ denote the normalised angular measure of $\bm{X}$ when its radial and angular components are defined with respect to some norm $\|\cdot\|$, that is
\begin{equation*}
\mathbb{P}(\bm{X}/\|\bm{X}\| \in \cdot \mid \|\bm{X}\| > t) \to H_{\bm{X}}(\cdot), \qquad (t\to\infty).
\end{equation*}
The goal is to find a (linear) subspace on which $H_{\bm{X}}$ is supported (or at least highly concentrated). Specifically, we assume that $H_{\bm{X}}$ is supported on a linear subspace $V^\star$ of dimension $p^\star < d$. 

The standard technique for identifying supporting (linear) subspaces, given iid observations of $\bm{X}$, is principal components analysis (PCA). If $\|\cdot\|=\|\cdot\|_1$, then $H_{\bm{X}}$ is a distribution on the unit simplex and correspondingly $\bm{X}/\|\bm{X}\|_1$ is a random composition. This opens up the possibility of employing CoDA-PCA for this task. On the other hand, taking $\|\cdot\|=\|\cdot\|_2$ results in a random vector $\bm{X}/\|\bm{X}\|_2$ and limiting distribution $H_{\bm{X}}$ that are not compositional, but rather circular/spherical. They are also restricted to the non-negative orthant, precluding the use of techniques fro the field of directional/spherical statistics [@fisherStatisticalAnalysisSpherical1987]. @dreesPrincipalComponentAnalysis2021 sweep this difficulty under the rug: they ignore the unit-norm constraint, and apply standard PCA to the the pseudo-angles as though they were points in $\R^d$. By grounding their method on Euclidean norms and distances they are able to derive statistical guarantees concerning reconstruction errors, i.e. the error incurred in representing an observation by its projection into a principal subspace. However, the utility of such guarantees may be called into question. First, for compositional data on the simplex, arbitrarily small Euclidean reconstruction errors can be arbitrarily large in the Aitchison metric. The discrepancy between the two metrics is most pronounced near the simplex boundary [@parkKernelMethodsRadial2022]. Accurate modelling of the angular measure in such regions is critical for risk assessment. Second, constrained data often exhibit curvature that cannot be described by standard linear dimension reduction techniques -- see Figure 1b in @aitchisonPrincipalComponentAnalysis1983 for an illustration of this phenomenon. Ultimately, this limits the dimension reducing capability of the method, since additional basis vectors are needed to reproduce the curvature. This leads us to suspect that, while the subspace detected by @dreesPrincipalComponentAnalysis2021 is optimal among a particular class of subspaces, it is not optimal with respect to a different (arguably more natural) class.

### CoDA-PCA for extremes

Fix $\|\cdot\|=\|\cdot\|_1$ and let $R:=\|\bm{X}\|_1$ and $\bm{\Theta}:=\bm{X}/\|\bm{X}\|_1$ denote the radial and angular components of $\bm{X}$. In this section, algebraic-geometric terms (e.g. linear, orthogonal, dimension, etc.) are to be interpreted in the Aitchison sense, unless stated otherwise. For any linear subspace $V\subset\mathbb{S}_+^{d-1}$, let $\Pi_V$ denote the orthogonal projection (matrix) onto $V$. In the spirit of @dreesPrincipalComponentAnalysis2021, we define the risk associated with $V$ by
\begin{equation*}
    R_{\infty}(V) := \mathbb{E}_{\bm{\Theta}\sim H_{\bm{X}}}[\| \bm{\Theta} \ominus \Pi_V \bm{\Theta} \|_a^2].
\end{equation*}
The risk $R_{\infty}(V)$ represents the expected squared reconstruction error under the limit model when $\bm{\Theta}$ is approximated by its projection $\Pi_V\bm{\Theta}$. Our working assumption is that there exists a $p^\star$-dimensional subspace $V^\star\subset\mathbb{S}_+^{d-1}$ such that $R_{\infty}(V^\star)=0$ and $R_{\infty}(V)>0$ for any subspace $V$ with $\mathrm{dim}(V)<p^\star$. In applications this assumption is only likely to hold approximately, introducing the familiar trade-off between dimension reduction and reconstruction error. Of course, the limit model is unknown and we cannot access samples from it, so it is impossible to compute $V^\star$ by minimising $R_{\infty}$ directly. Instead, we introduce the conditional risk of $V$ at a finite threshold $t>0$, defined by
\begin{equation*}
    R_t(V) := \mathbb{E}[ \| \bm{\Theta} \ominus \Pi_V \bm{\Theta} \|_a^2 \mid R > t].
\end{equation*}
Since the angular measure represents the limiting conditional distribution of angles above high thresholds, we intuitively expect that a minimiser $V_t^\star$ of $R_t$ is close to a minimiser of $R_\infty$, provided $t$ is sufficiently large. We estimate $V_t^\star$ by empirical risk minimisation. Let $\bm{X}_1,\ldots,\bm{X}_n$ be independent copies of $\bm{X}$ and define $R_i=\|\bm{X}_i\|_1$ and $\bm{\Theta}_i=\bm{X}_i/\|\bm{X}_i\|_1$ for $i=1,\ldots,n$. The empirical (conditional) risk is defined by
\begin{equation}\label{eq-pca-empirical-risk}
    \hat{R}_t(V) := \frac{1}{\sum_{i=1}^n \ind\{R_i > t\}} \sum_{i=1}^n \| \bm{\Theta}_i \ominus \Pi_V \bm{\Theta}_i \|_a^2 \ind\{R_i > t\}.
\end{equation}
The following result explains how to compute minimisers of these risk functions.

:::{#prp-coda-pca-extremes-risk-minimisers}
Fix $t>0$. Without loss of generality, assume that $\bm{\xi}_t := \mathrm{cen}_a(\bm{\Theta} \mid R >t) = \bm{e}_a$, i.e. $\bm{\Theta}$ is conditionally compositionally centred given $R>t$. Finally, assume that $\mathbb{E}[\|\bm{\Theta}\|_a^2 \mid R > t]<\infty$. 
\begin{enumerate}
    \item Let $\Sigma_t = \mathrm{Cov}(\mathrm{clr}(\bm{\Theta}) \mid R >t)$ be the conditional CLR-covariance matrix of $\bm{\Theta}$ given $R>t$. Suppose $\Sigma_t$ has eigenvalues $\lambda_1 \geq \lambda_2 \geq \ldots \geq \lambda_{d-1}\geq \lambda_d =0$ and corresponding eigenvectors $\bm{u}_1,\ldots,\bm{u}_d\in\R^d$. Then $V_p=\mathrm{span}_a(\{\mathrm{clr}^{-1}(\bm{u}_j) : j=1,\ldots,p\})$ minimises $R_t(V)$ among all linear subspaces $V$ of dimension $p$. It is the unique minimiser if $\lambda_p > \lambda_{p+1}$.
      \item Let $k=\sum_{i=1}^n \ind\{R_i > t\}$  and suppose the empirical conditional CLR-covariance matrix $\hat{\Sigma}_t = \frac{1}{k}\sum_{i=1}^k \mathrm{clr}(\bm{\Theta}_i)\mathrm{clr}(\bm{\Theta}_i)^T\ind\{R_i > t\}$ has eigenvalues $\hat{\lambda}_1 \geq \hat{\lambda}_2 \geq \ldots \geq \hat{\lambda}_{d-1}\geq \hat{\lambda}_d =0$ and corresponding eigenvectors $\hat{\bm{u}}_1,\ldots,\hat{\bm{u}}_d\in\R^d$. Then $\hat{V}_p=\mathrm{span}_a(\{\mathrm{clr}^{-1}(\hat{\bm{u}}_j) : j=1,\ldots,p\})$ minimises $\hat{R}_t(V)$ among all linear subspaces $V$ of dimension $p$. It is the unique minimiser if $\hat{\lambda}_p > \hat{\lambda}_{p+1}$.
\end{enumerate}
:::

::: {.proof}
Note that $\bm{Y}_t:=\mathrm{clr}(\bm{\Theta})\mid (R>t)$ is a random vector taking values in $\mathcal{T}^{d-1}\subset\R^d$ and 
\begin{equation*}
      \mathbb{E}[\|\bm{Y}_t\|_2^2] = \mathbb{E}[\|\mathrm{clr}(\bm{\Theta})\|_2^2\mid (R>t)] = \mathbb{E}[\|\bm{\Theta}\|_a^2 \mid R>t] < \infty,
\end{equation*}
by assumption. Thus $\bm{Y}_t$ satisfies the usual assumptions of unconstrained PCA in $\R^d$ and standard results, e.g. Theorem 5.3 in @seberMultivariateObservations1984, concerning minimiser(s) of $V\mapsto\mathbb{E}[\|\Pi_V \bm{Y}_t - \bm{Y}_t\|_2^2]$ yield (i). The argument follows analogously for (ii), except now $\bm{Y}_t$ following the empirical distribution of the CLR-transformed angular components associated with radial threshold exceedances.
:::

Minimising the expected angular reconstruction error is natural, but does not guarantee good performance in terms of estimation of the angular measure. Thus, we additionally consider the performance of the standard non-parametric estimator of the angular measure based on the compressed data versus the raw observations. For a given threshold $t>0$, the empirical angular measure is given by
\begin{equation}\label{eq-pca-empirical-angular-measure-full}
\hat{H}_{\bm{X}} = \frac{1}{\sum_{i=1}^n \ind\{R_i > t\}} \sum_{i=1}^n \delta_{\bm{\Theta}_i} \ind\{R_i > t\}.
\end{equation}
If the pseudo-angles are first projected onto a subspace $V$, then we obtain an alternative estimator
\begin{equation}\label{eq-pca-empirical-angular-measure-V}
\hat{H}_{\bm{X},V} = \frac{1}{\sum_{i=1}^n \ind\{R_i > t\}} \sum_{i=1}^n \delta_{\Pi_V\bm{\Theta}_i} \ind\{R_i > t\}.
\end{equation}
The probabilities associated with certain joint tail events can be expressed in terms of the angular measure, for example:
\begin{align}
\lim_{u\to\infty}\mathbb{P}(\min\bm{X} > u \mid \|\bm{X}\|_1 > u) &= \int_{\mathbb{S}_+^{d-1}}\left(\min_{j=1,\ldots,d}\theta_j\right)^\alpha \, H_{\bm{X}}(\dee\bm{\theta}), \label{eq-prob-min-H} \\
\lim_{u\to\infty}\mathbb{P}(\max\bm{X} > u \mid \|\bm{X}\|_1 > u) &= \int_{\mathbb{S}_+^{d-1}}\left(\max_{j=1,\ldots,d}\theta_j\right)^\alpha \, H_{\bm{X}}(\dee\bm{\theta}). \label{eq-prob-max-H}
\end{align}
These probabilities indicate how the components' contribution to extreme events are spread. If the underlying data-generating process is known, then the true values can be computed analytically or via Monte Carlo simulation. Replacing $H_{\bm{X}}$ with $\hat{H}_{\bm{X}}$ or $\hat{H}_{\bm{X},V}$ yields empirical estimates of these probabilities, the errors in which can be used to quantify the performance of our low-dimensional models. If the generative process is unknown, then we can still use the $\hat{H}_{\bm{X}}$-based estimate as a benchmark (based on all available information) against which to compare estimates obtained via $\hat{H}_{\bm{X},V}$.

### Simulation experiments

We now perform a series of simulation experiments comparing CoDA-PCA against the procedure of @dreesPrincipalComponentAnalysis2021, herein referred to as DS-PCA for brevity. 

One complication that arises is that CoDA-PCA and DS-PCA define angles with respect to different norms, so the sets of reconstructions are not immediately comparable. We resolve this by performing DS-PCA in the usual way, but projecting all points onto the simplex via self-normalisation with respect to $\|\cdot\|_1$ before computing the performance metrics. This ensures that the metrics are well-defined, except for DS-PCA projections that lie outside of the positive orthant. In such instances, the projected vector cannot properly be called a composition and the Aitchison reconstruction error is undefined. One might consider projecting it to the nearest point on the simplex, but this would lie on the boundary resulting in infinite Aitchison distances. In the absence of better options, we elect to discard such points. Arguably we are being charitable to DS-PCA by not directly penalising its tendency to produce invalid angles.

To guard against overfitting, we compute the empirical risk $\hat{R}_t$ across an unseen validation set, so that we are actually measuring out-of-sample reconstruction error. This ensures that the dimension reducing capabilities of the PCA model generalise to future observations. Specifically, for a fixed threshold $t>$, we detect the set of principal subspaces by applying @prp-coda-pca-extremes-risk-minimisers based on independent training samples $\bm{X}_1,\ldots,\bm{X}_n$, but compute the empirical risk \eqref{eq-pca-empirical-risk} on an unseen validation set $\bm{X}_1^\star,\ldots,\bm{X}_{n^\star}^\star$ of independent observations using the same threshold. The threshold $t$ is selected by specifying a desired number of extreme observations $k$ and setting $t:=R_{(k+1)}$, the $k+1$ order statistic of $\{R_1,\ldots,R_n\}$. Since we work in a simulation setting, the number of threshold exceedances in the validation set, roughly $n^\star k/n$, can be made arbitrarily large by increasing $n^\star$ accordingly.

#### Max-linear model with compositionally colinear factors

Our first experiment is based on the max-linear model with a parameter matrix that is carefully constructed to favour CoDA-PCA. This example is somewhat contrived, but illustrates the many benefits of our proposed methodology. To facilitate visualisation we restrict ourselves to $d=3$ dimensions, but the construction and our findings extend to higher dimensions. 

Let $\bm{u}^\star\in\mathbb{S}_+^{d-1}$ be a composition and suppose $A=(\bm{a}_1,\ldots,\bm{a}_q)\in\R_+^{d\times q}$ is such that, for all $j=1,\ldots,d$, there exists $\beta_j\in\R$ such that $\mathcal{C}\bm{a}_j = \beta_j \odot \bm{u}^\star$. In other words, the $q\geq 1$ columns of $A$ lie on the compositional straight line through $\bm{e}_a$ in the direction $\bm{u}^\star$. Let $\bm{Z}=(Z_1,\ldots,Z_q)$ with $Z_1,\ldots,Z_q$ independent standard Fréchet random variables. Suppose $\bm{X}=(X_1,\ldots,X_d)$ is defined by either of the following:
\begin{align}
\bm{X} &= A \circ \bm{Z} := \tau(A \tau^{-1}(\bm{Z})), \label{eq-maxlin-rv} \\
\bm{X} &= A \times_{\max} \bm{Z} := \left(\max_{j=1,\ldots,q}a_{1j}Z_j,\ldots,\max_{j=1,\ldots,q}a_{dj}Z_j\right), \label{eq-maxlin-ms}
\end{align}
where $\tau:\R\to(0,\infty)$ is the softplus function defined by
\begin{equation*}
\tau(y) := \log(1+\exp(y)).
\end{equation*}
Based on the discussion in Section 3 of @cooleyDecompositionsDependenceHighdimensional2019, we refer to random vectors generated by \eqref{eq-maxlin-rv} and \eqref{eq-maxlin-ms} as RV-max-linear and MS-max-linear, respectively. In either case, $\bm{X}$ is multivariate regularly varying with tail index $\alpha=1$ and common angular measure 
\begin{equation*}
    H_{\bm{X}}(\cdot) = \frac{\sum_{j=1}^q \|\beta_j \odot \bm{u}^\star\|_1 \delta_{\beta_j \odot \bm{u}^\star}(\cdot)}{\sum_{j=1}^q \|\beta_j \odot \bm{u}^\star\|_1} = \frac{1}{q}\sum_{j=1}^q \delta_{\beta_j \odot \bm{u}^\star}(\cdot).
\end{equation*}
This means that
\begin{equation*}
\mathrm{supp}(H_{\bm{X}}) = \{\beta_j \odot \bm{u}^\star : j=1,\ldots,q\} \subset \{\beta \odot \bm{u}^\star : \beta\in\R\}=\mathrm{span}_a(\bm{u}^\star),
\end{equation*}
a one-dimensional linear subspace of $\mathbb{S}_+^{d-1}$. The probabilities \eqref{eq-prob-min-H} and \eqref{eq-prob-max-H} are computed as
\begin{align}
\lim_{u\to\infty}\mathbb{P}(\min\bm{X} > u \mid \|\bm{X}\|_1 > u) &= \sum_{j=1}^q \min_{i=1,\ldots,d} a_{ij} \approx 0.636, \label{eq-prob-min-H-maxlinear-example}\\
\lim_{u\to\infty}\mathbb{P}(\max\bm{X} > u \mid \|\bm{X}\|_1 > u) &= \sum_{j=1}^q \max_{i=1,\ldots,d} a_{ij} \approx 0.0679. \label{eq-prob-max-H-maxlinear-example}
\end{align}
The reason for introducing the two generative processes is that they differ in terms of their finite sample properties, allowing a more detailed exploration of the finite-sample performance of our methodology. The angular components associated with large realisations of the MS-max-linear process tend to lie exactly at the discrete locations in $\mathcal{C}\bm{a}_j$, whereas for the RV-type process extremal angles tend to lie close to, but not exactly, at these points.

All simulations are based on $d=3$, $q=50$, $\bm{u}^\star=(0.12,0.58,0.3)$ and fixed values $\beta_1,\ldots\beta_{50}$ sampled uniformly between -4 and 4. We generate training sets of size $n\in\{5\times 10^3, 5\times 10^4\}$ and set $k/n\in\{1\%,5\%\}$. Reconstruction errors are based on validation sets of size $n^\star=n$. For each parameter combination, simulations are repreated 50 times.

```{r make-fig-sim-pca-example-data}
#| label: fig-sim-pca-example-data
#| fig-cap: "Examples of CoDA-PCA (left and middle) and DS-PCA (right) applied to RV- (left) and MS-max-linear (middle and right) data. The green diamonds represent the true angular measure. The black points are the angular components associated with the $k=50$ largest observations among a sample of size $n=5000$. The red and blue dashed lines represent the first and second principal axes, respectively. The red and blue crosses represent the projections onto the first and second principal subspaces, respectively."
#| fig-scap: "Example of CoDA-PCA and DS-PCA on max-linear data."
#| fig-height: 6
#| warning: false

A <- readRDS(file = file.path("scripts", "compositional", "results", "sim-pca-az-A-matrix.RDS"))
set.seed(12)
n <- 5000
k_frac <- 0.01

par(mfrow = c(1, 3), mar = c(0, 0, 0, 0), omi = c(0, 0, 0, 0))
# RV + coda pca
tmp <- extreme_pca(X_train = rmaxlinearfactor(n = n, A = A, type = "cooley"),
                   X_test = rmaxlinearfactor(n = n, A = A, type = "cooley"), 
                   k = floor(n * k_frac),
                   method = "coda", n_pcs = 1:2, plot = TRUE)
TernaryPoints(t(A), col = "green", cex = 1, pch = 5)
# generate MS data
X_tmp <- rmaxlinearfactor(n = n, A = A, type = "maxlin")
# coda pca
tmp <- extreme_pca(X_train = X_tmp, X_test = X_tmp, 
                   k = floor(n * k_frac), method = "coda", n_pcs = 1:2, plot = TRUE)
TernaryPoints(t(A), col = "green", cex = 1, pch = 5)

# ds pca
tmp <- extreme_pca(X_train = X_tmp, X_test = X_tmp,
                   k = floor(n * k_frac), method = "ds", n_pcs = 1:2, plot = TRUE)
TernaryPoints(t(A), col = "green", cex = 1, pch = 5)
```

Before diving in to the full simulation study, we invite the reader to examine @fig-sim-pca-example-data, which illustrates the example under consideration and will help provide some intuition for the results to follow. In each plot, the green diamonds represent the points $\mathcal{C}\bm{a}_j$ at which the angular measure places mass. These points follow a decidely curved pattern, but the essential structure is obviously one-dimensional. The black points represent the angular components of the threshold exceedances (here $k=50$ and $n=5000$). The left and middle ternary plots show the trends described by the first (red dashed line) and second (blue dashed line) compositional principal components. In the middle plot (MS-max-linear data), this red line follows the green points almost exactly, whereas the left-hand plot (RV-max-linear data) shows a degree of estimation error due to the weaker signal-to-noise ratio. The red crosses represent the rank-one reconstructions obtained by projecting the black points onto the red line. The second principal component describes all remaining variability in the data. In theory, one component is sufficient to describe the target distribution, but in finite sample settings an additional component is required (due to noise and the samples not coming directly from the limit model). The right-hand plot shows the results of DS-PCA applied to the MS-max-linear data. Now the first principal component is unable to capture the curvature of the data, yielding poor first-order projections (red crosses). (Note: these points lie on straight line in $\R^3$, but the process of visualising them on a ternary plot distorts the line slightly.) As discussed earlier, there are even a handful of projected points that lie outside of the ternary plot. Adding a second component improves the reconstructions significantly (blue crosses). Unlike CoDA-PCA, the two-dimensional approximations are still imperfect, because DS-PCA treats the angles as points in $\R^3$. 

With this example in mind, we consider the methods' performance over repeated simulations, shown in @fig-sim-pca-maxlin-loss. Within each sub-plot, we plot the empirical distribution of a given performance metric for each PCA method (bar colour) with varying number of principal components (bar outline). The maximum number of components is two, as including a third component does not add any information to the plot (*does this need explaining?*). The panels within each sub-plot correspond to the different combinations of $n$ and $k/n$. First, consider the Aitchison MSREs in the top-left plot. As expected, CoDA-PCA is able to reconstruct the data almost perfect with a single principal component. In contrast, the DS-PCA projections are relatively poor, even with the inclusion of a second component. This is primarily caused by imperfect reconstructions near the simplex boundary (see @fig-sim-pca-example-data, right), which are heavily penalised by the Aitchison metric. The top-right plot shows the Euclidean reconstruction error, i.e. \eqref{eq-pca-empirical-risk} with $\|\cdot\|_a$ replaced with $\|\cdot\|_2$. The main difference with the previous sub-plot is that now the two-dimensional DS-PCA are judged much more favourably. This shows how measuring performance using Euclidean distances can mask errors. Nevertheless, the CoDA method is vastly superior. This does not contradict the fact DS-PCA produces optimal subspaces [@dreesPrincipalComponentAnalysis2021, Lemma 2.1(iii)]. That optimality pertains to the class of linear subspaces in $\R^d$, which does not preclude the existence of better subspaces outside of this class. The bottom sub-plots show the empirical estimates of \eqref{eq-prob-min-H-maxlinear-example} and \eqref{eq-prob-max-H-maxlinear-example} obtained via the models \eqref{eq-pca-empirical-angular-measure-V}, where $V$ is the one/two principal subspace detected by each algorithm. With MS-max-linear data, 1D CoDA-PCA yields almost perfect estimates of both probabilities. When the data are generated from the RV-type process, the min and max probabilities tend to be overestimated/underestimated. This is because the sub-asymptotic distribution of the data is such that the first sample eigenvector $\hat{\bm{u}}_1$ is slightly biased for $\bm{u}^\star$. This is evidenced in @fig-sim-pca-example-data (left), where the red line gets pulled to the right of the green diamonds. Retaining a second component helps correct this, and the remaining bias in the probability estimates can be attributed to the noisy samples. Generally speaking, increasing the sample size reduces variance but does not eliminate biases. This means that the errors (or a lack thereof) can be attributed to the statistical methodology, rather than a lack of available data. Even with infinite samples, DS-PCA will fit a straight line to curved data!

```{r make-fig-sim-pca-maxlin-loss}
#| label: fig-sim-pca-maxlin-loss
#| fig-cap: "PCA performance metrics based on trivariate max-linear data."
#| fig-scap: "PCA performance metrics based on trivariate max-linear data."
#| fig-height: 6

res <- readRDS(file = file.path("scripts", "compositional", "results", "sim-pca-az.RDS"))

p1 <- res %>%
  mutate(maxlin_model = case_when(maxlin_model == "cooley" ~ "Regular variation",
                                  .default = "Max-stable")) %>%
  mutate(n_train = as.factor(n_train), k_frac = as.factor(label_percent()(k_frac))) %>%
  rename('n' = n_train, 'k/n' = k_frac) %>%
  pivot_longer(cols = c(test_loss_aitchison, test_loss_euclidean), names_to = "loss_type", values_to = "test_loss") %>%
  filter(loss_type == "test_loss_aitchison") %>%
  ggplot(aes(x = maxlin_model, y = test_loss, fill = pca_method, linetype = as.factor(n_pcs))) +
  geom_boxplot() +
  facet_grid(`k/n` ~ `n`, labeller = purrr::partial(label_both, sep = " = ")) +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.05))) +
  scale_fill_manual(labels = c("CoDA", "D&S"), values = c("red", "blue")) +
  scale_linetype_manual(labels = c("1", "2"), values = 1:2) +
  scale_x_discrete(labels = c("MS", "RV")) +
  labs(fill = "PCA method",
       linetype = "Number of PCs",
       x = "Generative model",
       y = "MSRE (Aitchison)") +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_blank()) +
  theme_light()

p2 <- res %>%
  mutate(maxlin_model = case_when(maxlin_model == "cooley" ~ "Regular variation",
                                  .default = "Max-stable")) %>%
  mutate(n_train = as.factor(n_train), k_frac = as.factor(label_percent()(k_frac))) %>%
  rename('n' = n_train, 'k/n' = k_frac) %>%
  pivot_longer(cols = c(test_loss_aitchison, test_loss_euclidean), names_to = "loss_type", values_to = "test_loss") %>%
  filter(loss_type == "test_loss_euclidean") %>%
  ggplot(aes(x = maxlin_model, y = test_loss, fill = pca_method, linetype = as.factor(n_pcs))) +
  geom_boxplot() +
  facet_grid(`k/n` ~ `n`, labeller = purrr::partial(label_both, sep = " = ")) +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.05))) +
  scale_fill_manual(labels = c("CoDA", "D&S"), values = c("red", "blue")) +
  scale_linetype_manual(labels = c("1", "2"), values = 1:2) +
  scale_x_discrete(labels = c("MS", "RV")) +
  labs(fill = "PCA method",
       linetype = "Number of PCs",
       x = "Generative model",
       y = "MSRE (Euclidean)") +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_blank()) +
  theme_light()

p3 <- res %>%
  mutate(maxlin_model = case_when(maxlin_model == "cooley" ~ "Regular variation",
                                  .default = "Max-stable")) %>%
  mutate(n_train = as.factor(n_train), k_frac = as.factor(label_percent()(k_frac))) %>%
  rename('n' = n_train, 'k/n' = k_frac) %>%
  ggplot(aes(x = maxlin_model, y = p_min_hat / 3, fill = pca_method, linetype = as.factor(n_pcs))) +
  geom_boxplot() +
  geom_hline(aes(yintercept = p_min / 3), colour = "black", linetype = "dashed") +
  facet_grid(`k/n` ~ `n`, labeller = purrr::partial(label_both, sep = " = ")) +
  scale_fill_manual(labels = c("CoDA", "D&S"), values = c("red", "blue")) +
  scale_linetype_manual(labels = c("1", "2"), values = 1:2) +
  scale_x_discrete(labels = c("MS", "RV")) +
  labs(fill = "PCA method",
       linetype = "Number of PCs",
       x = "Generative model",
       y = expression(lim(P(min(bold(X)) > u ~ "|" ~ "||"* bold(X) * "||"[1] > u), u %->% infinity))) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_blank()) +
  theme_light()

p4 <- res %>%
  mutate(maxlin_model = case_when(maxlin_model == "cooley" ~ "Regular variation",
                                  .default = "Max-stable")) %>%
  mutate(n_train = as.factor(n_train), k_frac = as.factor(label_percent()(k_frac))) %>%
  rename('n' = n_train, 'k/n' = k_frac) %>%
  ggplot(aes(x = maxlin_model, y = p_max_hat / 3, fill = pca_method, linetype = as.factor(n_pcs))) +
  geom_boxplot() +
  geom_hline(aes(yintercept = p_max / 3), colour = "black", linetype = "dashed") +
  facet_grid(`k/n` ~ `n`, labeller = purrr::partial(label_both, sep = " = ")) +
  scale_fill_manual(labels = c("CoDA", "D&S"), values = c("red", "blue")) +
  scale_linetype_manual(labels = c("1", "2"), values = 1:2) +
  scale_x_discrete(labels = c("MS", "RV")) +
  labs(fill = "PCA method",
       linetype = "Number of PCs",
       x = "Generative model",
       y = expression(lim(P(max(bold(X)) > u ~ "|" ~ "||"* bold(X) * "||"[1] > u), u %->% infinity))) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_blank()) +
  theme_light()

ggarrange(p1, p2, p3, p4, nrow = 2, ncol = 2, legend = "top", common.legend = TRUE)
```

#### Hüsler-Reiss model in low dimensions

Now we consider examples where the data are generated from a Hüsler-Reiss model. To start with we shall stay in three dimensions to facilitate visualisation. The relevance of this experiment is that, unlike the previous example, the true limit model is not designed to favour a particular PCA method.

Data are produced from three Hüsler-Reiss models parametrised by the following variograms (entries rounded to two decimal places):
\begin{equation*}
\Gamma_1 = 
\begin{pmatrix}
0.00 & 0.10 & 1.24 \\
0.10 & 0.00 & 0.67 \\
1.24 & 0.67 & 0.00
\end{pmatrix}, \qquad
\Gamma_2 = 
\begin{pmatrix}
0.00 & 0.14 & 0.04 \\
0.14 & 0.00 & 0.10 \\
0.04 & 0.10 & 0.00
\end{pmatrix}, \qquad
\Gamma_3 = 
\begin{pmatrix}
0.00 & 0.01 & 1.29 \\
0.01 & 0.00 & 1.24 \\
1.29 & 1.24 & 0.00
\end{pmatrix}.
\end{equation*}
These randomly generated variograms induce qualitatively different dependence structures, as shown in @fig-sim-pca-hr-threedim-example-data. The left plot ($\Gamma_1$) exhibits a curved trend with little variability in the direction orthogonal to this curve. This is a similar paradigm to the previous example. The extremes in the middle plot ($\Gamma_2$) is concentrated in near the barycentre of the simplex, implying strong asymptotic dependence between the three variables. The (empirical) angular measure exhibits a very slight curvature but is two-dimensional. Under the model parametrised by $\Gamma_3$ (right), extremes tend to occur in $X_1$ and $X_2$ jointly or $X_3$ singly. The extremal angles concentrate along a straight line joining the edge and vertex associated with these groups of components. 

```{r make-fig-sim-pca-hr-threedim-example-data}
#| label: fig-sim-pca-hr-threedim-example-data
#| fig-cap: "Example data from the three trivariate Hüsler-Reiss models. Based on $n=10^4$ and $k=250$."
#| fig-scap: "Example data from the three trivariate Hüsler-Reiss models."
#| fig-height: 5

Gamma_list <- readRDS(file = file.path("scripts", "compositional", "results", "sim-pca-hr-threedim-Gamma.RDS"))

par(mfrow = c(1, 3), mar = c(0, 0, 0, 0), oma = c(0, 0, 0, 0))
for (i in 1:3) {
  mev::rmev(n = 10000, d = 3, sigma = Gamma_list[[i]], model = "hr") %>%
    set_colnames(paste0("X", seq_len(ncol(.)))) %>%
    extreme_pca(X_train = .,
                X_test = ., 
                k = 250, 
                method = "coda", 
                n_pcs = 1:2, 
                plot = TRUE)
}
```

Similar to before, we repeatedly generate samples ($n=5000$) from each model and perform PCA based on the largest $k=50$ observations in norm. The probabilities \eqref{eq-prob-min-H} and \eqref{eq-prob-min-H} are computed empirically from samples of size $n=10^6$ using $u=100$. To three decimal places, the true values of \eqref{eq-prob-min-H} (resp. \eqref{eq-prob-max-H}) under the three sub-models are found to be 0.089, 0.228, 0.081 (resp. 0.603, 0.456, 0.586). For each simulated data set, we compute the MSRE and the error in the probability estimates obtained using low-rank reconstructions via \eqref{eq-pca-empirical-angular-measure-full} and \eqref{eq-pca-empirical-angular-measure-V}. The results are displayed in @fig-sim-pca-hr-threedim-loss. For $\Gamma_1$, one compositional principal component is sufficient to explain the data and produce good estimates. @dreesPrincipalComponentAnalysis2021 requires at least two components due to the non-linear trend. The data generated by $\Gamma_2$ is approximately linear and distinctly two-dimensional. Thus, both PCA procedures perform similarly and there is no scope for dimension reduction. The angular measure associated with $\Gamma_3$ concentrates along a straight line, which both methods are able to captured relatively well with a single eigenvector. The upshot is that CoDA-PCA performs at least as well as DS-PCA across a range of scenarios and outperforms it by a significant margin in some cases. Whether there is a difference in the methods depends on whether the low-dimensional target subspace $V^\star$ can be well-approximated by a linear subspace. In low dimensions this can be gauged by simply inspecting the data. Of course this is not generally feasible in high-dimensional applications, which represent the typical use case of such techniques. 
 
```{r make-fig-sim-pca-hr-threedim-loss}
#| label: fig-sim-pca-hr-threedim-loss
#| fig-cap: "PCA performance metrics based on trivariate Hüsler-Reiss data."
#| fig-scap: "PCA performance metrics based on trivariate Hüsler-Reiss data."
#| fig-height: 3.5

res <- readRDS(file = file.path("scripts", "compositional", "results", "sim-pca-hr-threedim.RDS")) 

p1 <- res %>%
  pivot_longer(cols = c(test_loss_aitchison, test_loss_euclidean), names_to = "loss_type", values_to = "test_loss") %>%
  filter(loss_type == "test_loss_aitchison") %>%
  ggplot(aes(x = as.factor(vario_seed_index), y = test_loss, fill = pca_method, linetype = as.factor(n_pcs))) +
  geom_boxplot() +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.05))) +
  scale_fill_manual(labels = c("CoDA", "D&S"), values = c("red", "blue")) +
  scale_linetype_manual(labels = c("1", "2"), values = 1:2) +
  scale_x_discrete(labels = c(expression(Gamma[1]), expression(Gamma[2]), expression(Gamma[3]))) +
  labs(fill = "PCA method",
       linetype = "Number of PCs",
       x = "Variogram",
       y = "MSRE (Aitchison)") +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_blank()) +
  theme_light()

p2 <- res %>%
  ggplot(aes(x = as.factor(vario_seed_index), y = p_min - (p_min_hat / 3), fill = pca_method, linetype = as.factor(n_pcs))) +
  geom_boxplot() +
  geom_hline(aes(yintercept = 0), colour = "black", linetype = "dashed") +
  scale_fill_manual(labels = c("CoDA", "D&S"), values = c("red", "blue")) +
  scale_x_discrete(labels = c(expression(Gamma[1]), expression(Gamma[2]), expression(Gamma[3]))) +
  scale_linetype_manual(labels = c("1", "2"), values = 1:2) +
  labs(fill = "PCA method",
       linetype = "Number of PCs",
       x = "Variogram",
       y = expression(lim(P(min(bold(X)) > u ~ "|" ~ "||"* bold(X) * "||"[1] > u), u %->% infinity) ~ "error")) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_blank()) +
  theme_light()

p3 <- res %>%
  ggplot(aes(x = as.factor(vario_seed_index), y = p_max - (p_max_hat / 3), fill = pca_method, linetype = as.factor(n_pcs))) +
  geom_boxplot() +
  geom_hline(aes(yintercept = 0), colour = "black", linetype = "dashed") +
  scale_fill_manual(labels = c("CoDA", "D&S"), values = c("red", "blue")) +
  scale_x_discrete(labels = c(expression(Gamma[1]), expression(Gamma[2]), expression(Gamma[3]))) +
  scale_linetype_manual(labels = c("1", "2"), values = 1:2) +
  labs(fill = "PCA method",
       linetype = "Number of PCs",
       x = "Variogram",
       y = expression(lim(P(max(bold(X)) > u ~ "|" ~ "||"* bold(X) * "||"[1] > u), u %->% infinity) ~ "error")) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_blank()) +
  theme_light()

ggarrange(p1, p2, p3, nrow = 1, ncol = 3, legend = "top", common.legend = TRUE)
```

#### Hüsler-Reiss model in high dimensions

For this experiment, the random vector $\bm{X}=(X_1,\ldots,X_{10})$ follows a Hüsler-Reiss distribution where the variogram $\Gamma\in\R_+^{10\times 10}$ is randomly generated according to the procedure of @fomichovSphericalClusteringDetection2023 (see Assumption A and Appendix B1 therein) with three clusters. The matrix of tail dependence coefficients associated with this variogram is given by $\chi_{ij}=2\bar{\Phi}(\sqrt{\Gamma_{ij}}/2)$. @fig-sim-pca-hr-highdim-param (left) provides a visual representation of this matrix. Recall that $\chi_{ij}=0$ if and only if $X_i$ and $X_j$ are asymptotically independent, and the magnitude of $\chi_{ij}$ indicates the extremal dependence strength between the corresponding pair of variables. We observe three groups/clusters and asymptotically dependent variables. Dependence is very strong among $\{X_1,\ldots,X_4\}$ and $\{X_9,X_{10}\}$, while the pairwise dependence strengths between the components in $\{X_5,\ldots,X_8\}$ is moderate and more variable. @fig-sim-pca-hr-highdim-param (right) shows the eigenvectors of the CLR-covariance matrix, estimated from a sample of size $n=10^6$ with $k=200$. The leading eigenvectors describe the extremal dependence structure with increasing resolution. The eigenvectors $\bm{u}_1,\bm{u}_2$ determine which cluster an extreme belong to, $\bm{u}_3,\bm{u}_4,\bm{u}_5$ capture the fine-scale behaviour within the second cluster. Subsequent eigenvectors account for patterns within the first and third clusters. We would speculate that retaining three to five principal components will be sufficient, depending on the complexity of the dependence in cluster two. This is borne out in @fig-sim-pca-hr-highdim-loss (top-right), which shows that first three components account for at least 95% of the total variability. In contrast, satisfying the same criterion (albeit with respect to a different notion of variance) under DS-PCA requires five components.  Compressing the data to three-dimensions yields no discernible deterioration in the probability estimates (bottom sub-plots). 

```{r make-fig-sim-pca-hr-highdim-param}
#| label: fig-sim-pca-hr-highdim-param
#| fig-cap: "Matrix of tail dependence coefficients $\\chi_{ij}=2\\bar{\\Phi}(\\sqrt{\\Gamma_{ij}}/2)$ (left) and a matrix of CoDA-PCA sample eigenvectors (right)."
#| fig-scap: "Tail dependence coefficients and CoDA-PCA eigenvectors for 10-dimensional Hüsler-Reiss model."
#| fig-height: 3.5

d <- 10
Gamma <- readRDS(file = file.path("scripts", "compositional", "results", "sim-pca-hr-highdim-Gamma.RDS"))
pc <- readRDS(file = file.path("scripts", "compositional", "results", "sim-pca-hr-highdim-loadings.RDS"))

# compute matrix of tail dependence coefficients
Chi <- 2 * (1 - pnorm(sqrt(Gamma) / 2)) 
colnames(Chi) <- paste0("X[", seq_len(ncol(Chi)), "]") %>% parse(text = .)

# compute matrix of eigenvectors
U_emp <- pc$loadings[seq_len(d), seq_len(d-1)]
rownames(U_emp) <- paste0("X[", seq_len(nrow(U_emp)), "]") %>% parse(text = .)
colnames(U_emp) <- paste0("bold(u)[", seq_len(ncol(U_emp)), "]") %>% parse(text = .)

# colour palette
n_colpal <- 13
colpal_seq <- colorspace::sequential_hcl(n = n_colpal, "Viridis")
colpal_div <- colorspace::diverging_hcl(n = n_colpal, "Blue-Red")

# plot Chi
plot_Chi <- t(Chi[rev(seq_len(nrow(Chi))), ])
p1 <- lattice::levelplot(plot_Chi, 
                         col.regions = colorspace::sequential_hcl(n = 13, "Viridis"),
                         at = lattice::do.breaks(c(0, 1), length(colpal_seq)),
                         xlab = "", ylab = "",
                         colorkey = TRUE,
                         scales = list(x = list(cex = 0.8, labels = parse(text = rownames(plot_Chi))),
                                       y = list(cex = 0.8, labels = parse(text = colnames(plot_Chi)))))
# plot loadings
plot_U_emp <- t(U_emp[rev(seq_len(nrow(U_emp))), ])
p2 <- lattice::levelplot(plot_U_emp, 
                         col.regions = colpal_div, 
                         at = lattice::do.breaks(c(-max(abs(U_emp)), max(abs(U_emp))), length(colpal_div)),
                         xlab = "", ylab = "",
                         colorkey = TRUE,
                         scales = list(x = list(cex = 0.8, labels = parse(text = rownames(plot_U_emp))),
                                       y = list(cex = 0.8, labels = parse(text = colnames(plot_U_emp)))))

ggarrange(p1, p2, ncol = 2)
```



```{r make-fig-sim-pca-hr-highdim-loss}
#| label: fig-sim-pca-hr-highdim-loss
#| fig-cap: "PCA performance metrics based on 10-dimensional Hüsler-Reiss data."
#| fig-scap: "PCA performance metrics based on 10-dimensional Hüsler-Reiss data."
#| fig-height: 6

res <- readRDS(file = file.path("scripts", "compositional", "results", "sim-pca-hr-highdim.RDS"))

p1 <- res %>%
  pivot_longer(cols = c(test_loss_aitchison, test_loss_euclidean), names_to = "loss_type", values_to = "test_loss") %>%
  filter(loss_type == "test_loss_euclidean") %>%
  ggplot(aes(x = as.factor(n_pcs), y = test_loss, fill = pca_method)) +
  geom_boxplot() +
  scale_fill_manual(labels = c("CoDA", "D&S"), values = c("red", "blue")) +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.05))) +
  labs(x = "Number of PCs",
       fill = "PCA method",
       y = "MSRE (Euclidean)") +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_blank()) +
  theme_light()

p2 <- res %>%
  filter(n_pcs <= 6) %>%
  group_by(n_pcs, pca_method) %>%
  summarise(prop_var = median(prop_var)) %>%
  ungroup() %>%
  ggplot(aes(x = n_pcs, y = prop_var, colour = pca_method)) +
  geom_point() +
  geom_line() +
  geom_hline(aes(yintercept = 0.95), colour = "black", linetype = "dashed") +
  scale_colour_manual(labels = c("CoDA", "D&S"), values = c("red", "blue")) +
  scale_y_continuous(limits = c(0, 1), expand = c(0, 0), labels = label_percent()) +
  labs(x = "Number of PCs",
       colour = "PCA method",
       y = "Cumulative prop. of (total) variance") +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_blank()) +
  theme_light()

p3 <- res %>%
  ggplot(aes(x = as.factor(n_pcs), y = p_min - (p_min_hat / 10), fill = pca_method)) +
  geom_boxplot() +
  geom_hline(aes(yintercept = 0), colour = "black", linetype = "dashed") +
  scale_fill_manual(labels = c("CoDA", "D&S"), values = c("red", "blue")) +
  labs(x = "Number of PCs",
       fill = "PCA method",
       y = expression(lim(P(min(bold(X)) > u ~ "|" ~ "||"* bold(X) * "||"[1] > u), u %->% infinity) ~ "error")) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_blank()) +
  theme_light()

p4 <- res %>%
  ggplot(aes(x = as.factor(n_pcs), y = p_max - (p_max_hat / 10), fill = pca_method)) +
  geom_boxplot() +
  geom_hline(aes(yintercept = 0), colour = "black", linetype = "dashed") +
  scale_fill_manual(labels = c("CoDA", "D&S"), values = c("red", "blue")) +
  labs(x = "Number of PCs",
       fill = "PCA method",
       y = expression(lim(P(max(bold(X)) > u ~ "|" ~ "||"* bold(X) * "||"[1] > u), u %->% infinity) ~ "error")) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_blank()) +
  theme_light()

ggarrange(p1, p2, p3, p4, nrow = 2, ncol = 2, legend = "top", common.legend = TRUE)
```


#### Max-linear model in high dimensions

- Same structure as earlier example, but now $d=10$, $q=25$, and $\mathcal{C}\bm{a}_j = \beta_{j1}\odot\bm{u}_1^\star \oplus \beta_{j2}\odot\bm{u}_2^\star$
- @fig-sim-pca-az-highdim-loss (left): Two/three components explain approx 95% of the total variance.
- @fig-sim-pca-az-highdim-loss (right): For MS data, two/three components gives good estimates. For RV data, four components gives good performance (corrects for error in previous eigenvectors). In each case, DS-PCA needs at least six components.


```{r make-fig-sim-pca-az-highdim-param}
#| label: fig-sim-pca-az-highdim-param
#| fig-cap: "Parameter matrix $A$ (left) and corresponding model TPDM (right) for 10-dimensional max-linear data."
#| fig-scap: "Parameter matrix $A$ and corresponding model TPDM for 10-dimensional max-linear data."
#| fig-height: 4

d <- 10

# A parameter matrix
A <- readRDS(file = file.path("scripts", "compositional", "results", "sim-pca-az-highdim-A.RDS"))
rownames(A) <- paste0("X[", seq_len(nrow(A)), "]") %>% parse(text = .)
colnames(A) <- paste0("bold(a)[", seq_len(ncol(A)), "]") %>% parse(text = .)

# TPDM
Sigma <- A^(1/2) %*% t(A^(1/2))
rownames(Sigma) <- paste0("X[", seq_len(nrow(Sigma)), "]") %>% parse(text = .)
colnames(Sigma) <- paste0("X[", seq_len(ncol(Sigma)), "]") %>% parse(text = .)

# colour palette
n_colpal <- 13
colpal_seq <- colorspace::sequential_hcl(n = n_colpal, "Viridis")
colpal_div <- colorspace::diverging_hcl(n = n_colpal, "Blue-Red")

# plot Chi
plot_A <- t(A[rev(seq_len(nrow(A))), ])
p1 <- lattice::levelplot(plot_A, 
                         col.regions = colorspace::sequential_hcl(n = 13, "Viridis"),
                         at = lattice::do.breaks(c(0, max(A)), length(colpal_seq)),
                         xlab = "", ylab = "",
                         colorkey = TRUE,
                         scales = list(x = list(draw = FALSE),
y = list(draw = FALSE))) %>%
  update(aspect = 0.6)
# plot loadings
plot_Sigma <- t(Sigma[rev(seq_len(nrow(Sigma))), ])
p2 <- lattice::levelplot(plot_Sigma, 
                         col.regions = colpal_seq, 
                         at = lattice::do.breaks(c(0, max(abs(Sigma))), length(colpal_div)),
                         xlab = "", ylab = "",
                         colorkey = TRUE,
                         scales = list(x = list(cex = 0.8, labels = parse(text = rownames(plot_Sigma))),
 y = list(cex = 0.8, labels = parse(text = colnames(plot_Sigma)))))

ggarrange(p1, p2, ncol = 2, widths = c(1, 0.8))
```

```{r make-fig-sim-pca-az-highdim-loss}
#| label: fig-sim-pca-az-highdim-loss
#| fig-cap: "PCA performance metrics based on 10-dimensional max-linear data."
#| fig-scap: "PCA performance metrics based on 10-dimensional max-linear data."
#| fig-height: 3

res <- readRDS(file = file.path("scripts", "compositional", "results", "sim-pca-az-highdim.RDS"))

p1 <- res %>%
  mutate(maxlin_model = case_when(maxlin_model == "cooley" ~ "Regular variation",
                                  .default = "Max-stable")) %>%
  filter(n_pcs <= 6) %>%
  group_by(n_pcs, pca_method, maxlin_model) %>%
  summarise(prop_var = median(prop_var)) %>%
  ungroup() %>%
  ggplot(aes(x = n_pcs, y = prop_var, colour = pca_method)) +
  geom_point() +
  geom_line() +
  geom_hline(aes(yintercept = 0.95), colour = "black", linetype = "dashed") +
  scale_colour_manual(labels = c("CoDA", "D&S"), values = c("red", "blue")) +
  scale_y_continuous(limits = c(0, 1), expand = c(0, 0), labels = label_percent()) +
  labs(x = "Number of PCs",
       colour = "PCA method",
       y = "Cumulative prop. of (total) variance") +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_blank()) +
  theme_light()

p2 <- res %>%
  mutate(maxlin_model = case_when(maxlin_model == "cooley" ~ "Regular variation",
                                  .default = "Max-stable")) %>%
  ggplot(aes(x = as.factor(n_pcs), y = p_max_hat / 10, fill = pca_method)) +
  geom_boxplot() +
  geom_hline(aes(yintercept = p_max / 10), colour = "black", linetype = "dashed") +
  facet_grid(~ maxlin_model) +
  scale_fill_manual(labels = c("CoDA", "D&S"), values = c("red", "blue")) +
  labs(fill = "PCA method",
       x = "Number of PCs",
       y = expression(lim(P(max(bold(X)) > u ~ "|" ~ "||"* bold(X) * "||"[1] > u), u %->% infinity))) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_blank()) +
  theme_light()

ggarrange(p1, p2, ncol = 2, legend = "top", common.legend = TRUE)
```

### Discussion

*Discuss conclusions of CoDA PCA stuff here.*

## Compositional classification for extremes

### Framework/motivation

Let $(\bm{X},Y)$ be a random pair with unknown joint distribution $F_{(\bm{X},Y)}$, where $Y\in\{-1,+1\}$ is a binary class label and $\bm{X}=(X_1,\ldots,X_d)$ is an $\R_+^d$-valued random vector containing covariate information that is presumed to be useful for predicting $Y$. For $\sigma\in\{-,+\}$, assume $\bm{X}\mid Y=\sigma 1$ is multivariate regularly varying with tail index $\alpha=1$ and angular measure $H_\sigma$ (with respect to a fixed norm $\|\cdot\|$ on $\R^d$). Given a labelled training sample $(\bm{X}_1,Y_1),\ldots,(\bm{X}_n,Y_n)$ of independent copies of $(\bm{X},Y)$, the goal is to find a classifier $g:\R^d\to\{-1,+1\}$ that minimises the expected classification error rate 
\begin{align*}
    \mathcal{L}(g) 
    &:= \mathbb{P}(Y\neq g(\bm{X})) \\
    &= \mathbb{P}(Y\neq g(\bm{X}) \mid \|\bm{X}\| \leq t) \mathbb{P}(\|\bm{X}\| \leq t) + \mathbb{P}(Y\neq g(\bm{X}) \mid \|\bm{X}\| > t) \mathbb{P}(\|\bm{X}\| > t).
\end{align*}
Since $F_{(\bm{X},Y)}$ is unknown, the standard approach to this task is estimate the statistical risk $\mathcal{L}(g)$ by the empirical risk
\begin{equation*}
    \hat{\mathcal{L}}(g)=(1/n)\sum_{i=1}^n \ind\{Y_i\neq g(\bm{X}_i)\}
\end{equation*}
and choose $\hat{g}=\argmin_{g\in\mathcal{G}}\hat{L}(g)$ over a suitable class $\mathcal{G}$. 
However, @jalalzaiBinaryClassificationExtreme2018 point out that the optimal classifier need not perform well in extreme regions of the predictor space, e.g. $\{\|\bm{X}\|>t\}$ for $t>0$ large, since such regions exert a negligible influence over the global prediction error. This motivates the idea of building a classifier that minimises the asymptotic risk in the extremes, defined as
\begin{equation*}
    \mathcal{L}_{\infty}(g) := \lim_{t\to\infty}\mathbb{P}(Y\neq g(\bm{X}) \mid \|\bm{X}\| > t).
\end{equation*}
They prove that, under certain assumptions, the minimiser $g^\star := \argmin_{g\in\mathcal{G}}\mathcal{L}_{\infty}(g)$ is of the form $g^\star(\bm{x})=g^\star(\bm{x}/\|\bm{x}\|)$ [@jalalzaiBinaryClassificationExtreme2018, Theorem 1]. In practice, this suggests finding solutions of the minimisation problem
\begin{equation}\label{eq-extreme-classification-optimisation}
    \min_{g\in\mathcal{G}_{\bm{\Theta}}}\hat{L}_t(g),\qquad \hat{\mathcal{L}}_t(g) = \frac{1}{\sum_{i=1}^n \ind\{\|\bm{X}\| > t\}} \sum_{i=1}^n \ind\left\lbrace Y_i \neq g\left(\frac{\bm{X}_i}{\|\bm{X}_i\| }\right), \|\bm{X}\| >t\right\rbrace.
\end{equation}
for some high threshold $t>0$, where $\mathcal{G}_{\bm{\Theta}}$ denotes a family of classifiers $g:\mathbb{S}_+^{d-1}\to\{-1+1\}$. 

The remainder of their paper is devoted to providing theoretical guarantees for this learning principle, leaving aside "the practical issue of designing efficient algorithms for solving \eqref{eq-extreme-classification-optimisation}". This is exemplified in their numerical experiments, where they simply resort to popular, general-purpose classifiers such as $k$-NN and random forests. These algorithms disregard the unit-norm constraint imposed on the input data. By now, we hope the reader is persuaded that this mismatch can have significant practical ramifications. Upon taking $\|\cdot\|=\|\cdot\|_1$, \eqref{eq-extreme-classification-optimisation} becomes a compositional binary classification problem. The CoDA community develops bespoke algorithms for such tasks. Implementations of these algorithms are readily available in packages such as \texttt{Compositional} and \texttt{CompositionalML}.

### Simulation experiments

For our simulation experiments, we generate realisations of $\bm{X}\mid Y=y$ on standard Fréchet margins from one of three MEV models: symmetric logistic, asymmetric logistic, and bilogistic. The negative ($y=-1$) and positive ($y=+1$) class instances are generated using different (scalar) dependence parameters, denoted $\vartheta_0$ and $\vartheta_1$, respectively. The classes are balanced globally and asymptotically, meaning
\begin{equation*}
p=\mathbb{P}(Y=+1)=0.5, \qquad p_{\infty}:=\lim_{t\to\infty}\mathbb{P}(Y=+1\mid\|\bm{X}\|_1>t)=0.5.
\end{equation*}
From each model, we simulate a labelled training set $(\bm{x}_1,y_1),\ldots,(\bm{x}_n,y_n)$ of size $n=5\times 10^3$. Each tail classifier is trained using the $k=500$ largest observations (in $L_1$-norm) among this set. @fig-sim-classification-ternary illustrates one realisation of this subset for each model. Each plot gives an indication of the two classes' tail dependence structures and the degree of difficulty in classifying them. The performance of each classifier will be assessed by its asymptotic risk. This is estimated empirically using a validation set comprising $10^5$ samples from the limit model, i.e. angles sampled from the angular measures $H_{-}$ and $H_{+}$ using the $\texttt{rmevspec}$ function from the $\texttt{mev}$ package. In practical scenarios where we cannot access unlimited samples from the limit model, we would instead assess the extrapolation capacity of the classifier by compute the empirical risk at a sequence of increasing thresholds on a hold-out validation set. All reported results are based on 100 repeated simulations. 

```{r make-fig-sim-classification-ternary}
#| label: fig-sim-classification-ternary
#| fig-cap: "Blah."
#| fig-scap: "Blah."
#| fig-height: 5
#| message: false

library(ggtern)
readRDS(file = file.path("scripts", "compositional", "results", "classification-sim-ternary-data.RDS")) %>%
  mutate(nice_model = paste0(nice_model, ": " , "vartheta[0] == ", param0)) %>%
  mutate(param1_frac = paste0("vartheta[1]/vartheta[0] == ", param1_frac)) %>%
  ggtern(aes(X1, X2, X3, colour = class)) +
  geom_mask() +
  geom_point(shape = 4) +
  facet_grid(param1_frac ~ nice_model, labeller = label_parsed) +
  scale_colour_manual(values = c("red", "blue"), labels = c(expression(vartheta[0]), expression(vartheta[1]))) +
  xlab("") +
  ylab("") +
  labs(colour = "Class dependence parameter") +
  Llab(expression(X[1])) +
  Tlab(expression(X[2])) +
  Rlab(expression(X[3])) +
  theme_hidegrid() +
  theme_hidelabels() +
  theme(legend.position = "bottom")
detach(package:ggtern, unload = TRUE)
```

The catalogue of classification algorithms is virtually limitless. For simplicity, we restrict ourselves to three types: $k$-nearest neighbours, support vector machines, and random forests. Since our primary goal is to compare CoDA-based classifiers to standard ones, we shall employ the $\alpha$-transformed classifiers described earlier. For fixed $\alpha\in\{0, 0.1, \ldots, 0.9, 1\}$, the classifiers' tuning parameters are selected by ten-fold cross-validation on the training set. The hyperparameters of the three classifiers considered here are described in XXX(A). The tuning procedure is summarised as follows:

1. Let $\mathcal{G}_{\bm{\Theta}}=\{g_{\psi}:\psi\in\Psi\}$ be a family of simplicial classifiers $g_\psi:\mathbb{S}_+^{d-1}\to\{-1,+1\}$ parametrised by $\psi\in\Psi$. For example, if $\mathcal{G}_{\bm{\Theta}}$ represents the $k$-NN($\alpha$) class with fixed $\alpha$, then $\psi=k$ and $\Psi=\mathbb{N}$. 
2. For a given training set $\{(\bm{x}_i, y_i) : i=1,\ldots,n\}$, let $\bm{\theta}_{(1)},\ldots,\bm{\theta}_{(k)}$ be the angular components of the $k$ largest observations and $y_{(1)},\ldots,y_{(k)}$ their associated classes. Let $t$ denote the implicit threshold, i.e. the $k+1$ order statistic of $\{\|\bm{x}_i\|_1 : i=1,\ldots,n\}$
3. Partition $\bm{\theta}_{(1)},\ldots,\bm{\theta}_{(k)}$ into $J=10$ balanced folds by stratified sampling. 
4. For $j=1,\ldots,J$ and $\psi\in\Psi$, predict the classes of the elements of fold $j$ by applying the classification rule $g_{\psi}$ trained on all the (labelled) data not in fold $j$. Denote the resulting classification error rate by $\hat{\mathcal{L}}_t^{(j)}(g_{\psi})$. 
5. Estimate the risk of $g_{\psi}$ at level $t$ as $$\hat{\mathcal{L}}_t(g_{\psi}) = \frac{1}{J}\sum_{j=1}^J \hat{\mathcal{L}}_t^{(j)}(g_{\psi}).$$
6. Select $\psi$ to minimise the empirical risk, that is $\hat{\psi}=\arg\min_{\psi\in\Psi} \hat{\mathcal{L}}_t(g_{\psi})$. The tuned classifier is $\hat{g}=g_{\hat{\psi}}$.
7. Compute the asymptotic risk $\mathcal{L}_{\infty}(\hat{g})$ of $\hat{g}$ as the empirical classification error rate based on the samples from the true limit model. By taking sufficiently many Monte Carlo samples, the asymptotic risk can be computed to arbitrary precision.  

@fig-sim-classification-asymptotic-risk presents the results of our experiment. Each sub-panel corresponds to a generative process, that is, the MEV model and the ratio between the dependence parameters. Within each sub-panel, we plot the asymptotic risk as a function of the data-transformation parameter $\alpha$. The solid lines represent the median risk (across the 100 repeats) while the bands depict the interquartile range. The colours indicate the classifier type. The ratio of the dependence parameters dictates the difficulty level of the learning task. The error rates are between 30-40% when $\vartheta_1/\vartheta_0=1.5$ and between 2-16% when $\vartheta_1/\vartheta_0=3$. Universally, the statistical risk is maximised when the underlying geometry is Euclidean ($\alpha=1$). For the bilogistic and symmetric logistic models, $\alpha=0$ appears optimal. For the negative logistic data, the minimal risk is attained at some intermediate value, say $\alpha\approx 0.3$. Thus the optimal classifiers fall somewhere under the CoDA umbrella, corresponding to either  the Aitchison metric ("quintessential CoDA") or the $\alpha$-metric with $\alpha\neq 1$ ("modern CoDA"), respectively. The choice of classifier is obviously a key determinant of performance, with $k$-NN($\alpha$) typically being the worst-performing and $\alpha$-SVM the best. Notwithstanding this, we highlight that $k$-NN($\alpha=0$) is usually fairly competitive against the Euclidean SVM/RF. This shows that a simple classifier in the 'correct' geometry can be as good as a sophisticated classifier in the 'wrong' geometry.

```{r make-fig-sim-classification-asymptotic-risk}
#| label: fig-sim-classification-asymptotic-risk
#| fig-cap: "Blah."
#| fig-scap: "Blah."
#| fig-height: 6

readRDS(file = file.path("scripts", "compositional", "results", "classification-sim-results-knn-svm-rf.RDS")) %>%
  mutate(classifier = case_when(classifier == "knn" ~ 'k * "-NN" * (alpha)',
                                classifier == "svm" ~ 'alpha * "-SVM"',
                                classifier == "rf" ~ 'alpha * "-RF"')) %>%
  mutate(param0 = paste0("vartheta[0] == ", param0),
         param1_frac = paste0("vartheta[1]/vartheta[0] == ", param1_frac)) %>%
  ggplot(aes(x = alpha, y = test_error, colour = classifier, fill = classifier)) +
  stat_summary(geom = "line", fun = median, linewidth = 0.7) +
  stat_summary(geom = "ribbon", fun.min = function(x) quantile(x, 0.25), fun.max = function(x) quantile(x, 0.75), alpha = 0.1, colour = NA) +
  facet_nested_wrap(param1_frac ~ nice_model, scales = "free", labeller = label_parsed, nrow = 2,
                    nest_line = element_line(colour = "white")) +
  scale_x_continuous(limits = c(0, 1), expand = c(0, 0), breaks = breaks_pretty(n = 5)) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.05)), breaks = breaks_pretty(n = 4), labels = label_percent()) +
  scale_colour_manual(values = c("red", "blue", "darkgreen"), labels = parse_format()) +
  scale_fill_manual(values = c("red", "blue", "darkgreen"), labels = parse_format()) +
  xlab(expression(alpha)) +
  ylab("Asymptotic risk") +
  labs(fill = "Classifier", colour = "Classifier") +
  theme_light() +
  theme(legend.position = "top",
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank())
```

In practice, the asymptotic risk cannot form the basis for our choice of $\alpha$, since it is a non-observable quantity. Then, $\alpha$ becomes an additional hyperparameter in the model, which may be selected by cross-validation along with the other tuning parameters. The results of this approach are presented in @fig-sim-classification-cv-alpha. Roughly speaking, the selected values of $\alpha$ accord with our earlier findings, with $\alpha\approx 0.3$ being favoured in the negative logistic case and $\alpha=0$ in the other two. However, there is significant variation in the selected values. Moreover, $\alpha=1$ is chosen in a non-negligible proportion of runs, despite this being the worst choice according to the asymptotic risk criterion. This suggests that $\hat{L}_t(\cdot)$ and $\hat{L}_{\infty}(\cdot)$, when viewed as functions of $\alpha$, can exhibit differing profiles. 


```{r make-fig-sim-classification-cv-alpha}
#| label: fig-sim-classification-cv-alpha
#| fig-cap: "Blah."
#| fig-scap: "Blah."
#| fig-height: 6

readRDS(file = file.path("scripts", "compositional", "results", "classification-sim-results-knn-svm-rf.RDS")) %>%
  mutate(classifier = case_when(classifier == "knn" ~ 'k * "-NN" * (alpha)',
                                classifier == "svm" ~ 'alpha * "-SVM"',
                                classifier == "rf" ~ 'alpha * "-RF"')) %>%
  mutate(param0 = paste0("vartheta[0] == ", param0),
         param1_frac = paste0("vartheta[1]/vartheta[0] == ", param1_frac)) %>%
  group_by(nice_model, param0, param1_frac, classifier, rep) %>%
  slice_min(train_error, n = 1) %>%
  ggplot(aes(x = alpha, fill = classifier)) +
  geom_density(alpha = 0.7) +
  facet_nested_wrap(param1_frac ~ nice_model, scales = "free", labeller = label_parsed, nrow = 2,
                    nest_line = element_line(colour = "white")) +
  scale_x_continuous(limits = c(0, 1), expand = c(0,0), breaks = breaks_pretty(n = 5)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05)), breaks = breaks_pretty(n = 4)) +
  scale_fill_manual(values = c("red", "blue", "darkgreen"), labels = parse_format()) +
  xlab(expression(alpha)) +
  ylab("Asymptotic classification risk") +
  labs(fill = "Classifier") +
  theme_light() +
  theme(legend.position = "top", 
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank()) 
```

This hypothesis is borne out by @fig-sim-classification-train-test-risk, which plots the median values of the 'training loss' $\hat{L}_t$ (dashed lines) and 'test loss' $\hat{L}_{\infty}$ (solid lines) against $\alpha$. Indeed, for the negative logistic model with $\vartheta_1/\vartheta_0=3$, the training loss of $k$-NN($\alpha$) is almost flat, while the test loss exhibits a definite positive gradient. This explains why the optimal $\alpha$ values are almost uniformly distributed -- see the green kernel density estimate in bottom-middle panel of @fig-sim-classification-cv-alpha. *Speculate as to what is going on here, and give some concluding remarks. Influence of $n$ and $k$? More work needed on estimating asymptotic risk?* 

```{r make-fig-sim-classification-train-test-risk}
#| label: fig-sim-classification-train-test-risk
#| fig-cap: "Blah."
#| fig-scap: "Blah."
#| fig-height: 6

readRDS(file = file.path("scripts", "compositional", "results", "classification-sim-results-knn-svm-rf.RDS")) %>%
  mutate(classifier = case_when(classifier == "knn" ~ 'k * "-NN" * (alpha)',
                                classifier == "svm" ~ 'alpha * "-SVM"',
                                classifier == "rf" ~ 'alpha * "-RF"')) %>%
  mutate(param0 = paste0("vartheta[0] == ", param0),
         param1_frac = paste0("vartheta[1]/vartheta[0] == ", param1_frac)) %>%
  pivot_longer(cols = c("train_error", "test_error"), names_to = "error_type", values_to = "error") %>%
  ggplot(aes(x = alpha, y = error, colour = classifier,  linetype = error_type)) +
  stat_summary(geom = "line", fun = median) +
  facet_nested_wrap(param1_frac ~ nice_model, scales = "free", labeller = label_parsed, nrow = 2,
                    nest_line = element_line(colour = "white")) +
  scale_x_continuous(limits = c(0, 1), expand = c(0.01, 0.01), breaks = breaks_pretty(n = 5)) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.05)), breaks = breaks_pretty(n = 4), labels = label_percent()) +
  scale_colour_manual(values = c("red", "blue", "darkgreen"), labels = parse_format()) +
  scale_fill_manual(values = c("red", "blue", "darkgreen"), labels = parse_format()) +
  scale_linetype_manual(values = 1:2, labels = c("Asymptotic (test set)", "Empirical at level t (train set)")) +
  xlab(expression(alpha)) +
  ylab("Risk") +
  labs(linetype = "Risk type", colour = "Classifier") +
  theme_light() +
  theme(legend.position = "top",
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank())
```

### Discussion

*Discuss conclusions of CoDA classification stuff here.*

## Appendix material

### (A) Computational details regarding the $k$-NN($\alpha$), $\alpha$-SVM and $\alpha$-RF classifiers

*List the hyperparameters of each method, describe what they mean, list the ranges of values used, and give any relevant computational details. Refer to $\texttt{Compositional}$ and $\texttt{CompositionalML}$ packages.*







